### 2013-05-14 paper.tex
First version

\documentclass{article}
\begin{document}

\title {Why robot programming is hard}
\author {Tessa Lau \\
Willow Garage \\
Menlo Park, CA USA \\
{\tt tlau@willowgarage.com}
}
\maketitle

Why robotics is becoming important
Statistics on size of market
Number of research investments?
Roomba, Neato in consumer use
	http://www.neatorobotics.com/
	http://store.irobot.com/home/index.jsp
Baxter human-safe 2-armed manipulator
	runs ROS
	http://www.ros.org/news/2012/09/rethink-ros.html
	http://www.rethinkrobotics.com/index.php/products/baxter/
Industrial robots:
	KUKA robot arm http://en.wikipedia.org/wiki/KUKA
Toyota's human support robot:
	runs ROS
	http://www.gizmag.com/toyota-human-support-robot/24246/
Telepresence robots:
	Beam
	Anybot QB
	Double

Main points:
	EUP is what will enable robotics to spawn new industries
	by enabling businesses to customize robot behavior for their own needs

	Yet programming robots is difficult
	State of the art today is demonstrational, which works well for limited, repetitive tasks
	Or using highly technical tools to specify fine-grained robot behavior
	Behavior specification requires much deep technical knowledge
	Not only the traditionally hard topics in EUP such as conditionals, loops, variables
	But also advanced topics like concurrency, uncertainty, high-dimensionality

\section{ROS}

History of ROS
Invented by Willow Garage, new operating system for diverse robots
Current reach
running on robots from PR2 to android-powered XXX
Provides very rich control substrate for programming the behavior of complex robots

\section{Concurrency}

ROS nodes
multiple parts of robots (head, arms, base)
callbacks, coroutines, threads

\section{Uncertainty}

World changes out from under you
Hardware fails
Incomplete world models
Bayesian methods, probabilities

\section{High-dimensionality}

Robot arms have 7DOF
grasps require 6D
world is 3D
quaternions, matrix manipulation
specialized I/O devices

\bibliographystyle{plain}
\bibliography{general}

\end{document}

### 2013-05-14 paper.tex
Work work

@@
 Number of research investments?
+
+Main points:
+	EUP is what will enable robotics to spawn new industries
+	by enabling businesses to customize robot behavior for their own needs
+
+	Yet programming robots is difficult
+	State of the art today is demonstrational, which works well for limited, repetitive tasks
+	Or using highly technical tools to specify fine-grained robot behavior
+	Behavior specification requires much deep technical knowledge
+	Not only the traditionally hard topics in EUP such as conditionals, loops, variables
+	But also advanced topics like concurrency, uncertainty, high-dimensionality
+
+The challenge for robotics is to bring robots to market that are consumer-friendly and can operate in human environments.
+For limited tasks, we are seeing robots that can perform one single task very well without getting in the way of the humans in the household.
+Vacuum cleaner robots like the iRobot's Roomba and the Neato XV are flooding the market.
+The Robomow robotic lawn mower performs a similar task for your lawn.
+Undoubtedly their success is due to their extremely simple interface: the Roomba has a single large button labeled ``Clean''.
+
+\section{Related work}
+
 Roomba, Neato in consumer use
@@
 	Double
-
-Main points:
-	EUP is what will enable robotics to spawn new industries
-	by enabling businesses to customize robot behavior for their own needs
-
-	Yet programming robots is difficult
-	State of the art today is demonstrational, which works well for limited, repetitive tasks
-	Or using highly technical tools to specify fine-grained robot behavior
-	Behavior specification requires much deep technical knowledge
-	Not only the traditionally hard topics in EUP such as conditionals, loops, variables
-	But also advanced topics like concurrency, uncertainty, high-dimensionality
 

### 2013-05-17 paper.tex
More work

@@
 
-\title {Why robot programming is hard}
+\title {Robot programming for dummies}
 \author {Tessa Lau \\
@@
 \maketitle
+
+I never thought I would be a roboticist.
+In grad school, our robotics team spent months teaching little robot dogs to play soccer and solving other seemingly simple problems.
+I was better off in the world of software, I thought, where the systems I built could make a real difference in people's lives.
+Software robots would automate repetitive tasks in computer use and free people from mundane work.
+End user programming would enable millions of ordinary computer users to customize and adapt systems to their own needs.
+
+But that world is changing.
+People aren't primarily sitting in front of their computers, pushing bits around.
+Nowadays, innovation is happening in the real world, where computing is being used to effect physical change in the world around us.
+The past decades have seen industrial robots increase manufacturing efficiency on the factory floor.
+Even more, we are now seeing robots popping up in the consumer space.
+Robotic vacuum cleaners are changing the way we clean our homes.
+Robotic lawnmowers keep the yard tidy without having to lift a finger.
+Nest's robotic thermostat learns your habits and keeps your home at the optimal temperature.
+
+All these innovations have become possible through multiple advances: in sensor technology, which robots use to perceive the world around them; in planning and navigation, which enable robots to move around in the world unaided; in human-safe motion controllers, which let robots move their limbs without endangering humans around them; in the ROS open source robot operating system, which enables roboticists to build on each others' work without starting from scratch; and many more advances in the broader field of robotics.
+
+Yet despite all of these advances, the holy grail of personal robotics remains elusive.
+When will we have our own Rosie the Robot (from the Jetsons cartoon) to take over housekeeping chores?
+Since starting work at Willow Garage and becoming immersed in the challenges of the robotics industry, I now believe that the limiting factor is end user programming.
 

### 2013-05-20 paper.tex
More word hacking

--- 2013-05-14 paper.tex
@@
 \maketitle
+
+I never thought I would be a roboticist.
+In grad school, our robotics team spent months teaching little robot dogs to play soccer and solving other seemingly simple problems.
+Far better to work in the world of software, I figured, where the systems I built could make a real difference in people's lives.
+Software robots would automate repetitive tasks in computer use and free people from mundane work.
+End user programming would enable millions of ordinary computer users to customize and adapt systems to their own needs.
+
+But that world is changing.
+Computing has moved off the desktop and into the smartphones, smart environments, and the world aronud us.
+Innovation is happening in the real world, using computing to effect physical change in our environment.
+The past decades have seen industrial robots increase manufacturing efficiency on the factory floor.
+Even more, we are now seeing robots tackle tasks from our daily lives.
+Robotic vacuum cleaners are changing the way we clean our homes.
+Robotic lawnmowers keep the yard tidy without having to lift a finger.
+Nest's robotic thermostat learns your habits and keeps your home at the optimal temperature.
+Google's driverless car could change the way we commute.
+
+All these innovations have become possible through multiple advances: in sensor technology, which robots use to perceive the world around them; in planning and navigation, which enable robots to move around in the world unaided; in human-safe motion controllers, which let robots move their limbs without endangering the people around them; in the ROS open source robot operating system, which enables roboticists to build on each others' work without starting from scratch; and many more advances in the broader field of robotics.
+
+Yet the holy grail of personal robotics remains elusive.
+When will we have our own Rosie the Robot (from the Jetsons cartoon) to take over housekeeping chores?
+When will disabled adults have robot helpers to assist them in activities of daily living?
+All of these advances in robotics still fall short of the goal of creating personal robots that can do whatever we want them to do.
+
+Actually, the problem is not that robots {\em can't} do what we want them to do.
+The problem is that it's too difficult to {\em tell} robots what we want them to do.
+The first company to crack the nut of end user robot programming will corner the market on household robots.
+Human-robot interaction is the missing link for widespread personal robotics.
+
+Let's look at the state of the art in robot programming technology.
+
+\section{ROS}
+
+% History of ROS
+% Invented by Willow Garage, new operating system for diverse robots
+% Current reach
+% Running on robots from PR2 to android-powered XXX
+% Provides very rich control substrate for programming the behavior of complex robots
+
+The Robot Operating System is one of the software advances that supports the rapid development of new robotics technology.
+First developed in 2007 at the Stanford Articifial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
+
+ROS provides a distributed publish/subscribe messaging platform for a collection of {\em nodes} to communicate with each other.
+
+Before ROS, roboticists had to create the software to control their robots from scratch, each doing the same work to create device drivers and rebuild libraries for their specific robot's hardware.
 
@@
 
+read this: http://workshop.iroboticist.com/why-robotics/
+
 \section{Related work}
@@
 	Double
-
-\section{ROS}
-
-History of ROS
-Invented by Willow Garage, new operating system for diverse robots
-Current reach
-running on robots from PR2 to android-powered XXX
-Provides very rich control substrate for programming the behavior of complex robots
 
@@
 
+\section{Speech interfaces}
+
+Impedance mismatch of speech interfaces
+
 \bibliographystyle{plain}

### 2013-05-28 paper.tex
More writing

@@
 
-Let's look at the state of the art in robot programming technology.
+\section{Case study: Robots for Humanity}
+
+One potential use case for personal robotics will be in-home care and assistance for disabled adults.
+The Robots for Humanity project, a collaboration between Willow Garage and Georgia Tech, aims to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
+Its first prototypes have enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
+Limited to the ability to move a single finger, Henry has been able to instruct Willow Garage's PR2 robots to open cabinets, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+
+Yet our experience with Robots for Humanity underscores how difficult it is for ordinary mortals to instruct robots what to do.
+Even using state-of-the-art robot visualization software developed at Willow Garage, simple tasks such as opening a cabinet door could take Henry upwards of 15 minutes to accomplish.
+While Henry has only limited interaction capabilities due to motor impairment, he also has incredible motivation.
+Ordinary consumers, on the other hand, while fully able-bodied, may not share the same perseverance in programming robots.
+
+Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
+Why is end user robot programming so challenging?
+Several compounding factors make it more difficult than normal computer use.
+
+\section{High dimensionality}
+
+% Robot arms have 7DOF
+% grasps require 6D
+% world is 3D
+% quaternions, matrix manipulation
+% specialized I/O devices
+
+Unlike traditional computer interfaces which are restricted to two dimensions, robot control requires manipulating objects in a three-dimensional world.
+Visualizing a 3D world on existing 2D displays requires the extra dimension to be projected down into two dimensions, giving rise to the popular ``first person shooter'' style of interface.
+These interfaces display the world as perceived from a virtual camera, where objects can be occluded by other objects.
+Additional camera controls are needed to pan/zoom around the space to overcome occlusion, and these controls can be nonintuitive to non-expert computer users.
+
+High dimensionality is even more important in specifying how to grasp and place objects in the world.
+A solid object has six degrees of freedom: three degrees to specify its position, and another three to specify its rotation.
+The human arm, and the most advanced robotic arms, have seven degrees of freedom.
+Many robot arm control interfaces ask users to independently specify joint angles for each of those seven degrees of freedom.
+These interfaces are cumbersome because the joints are interdependent: setting the angle of one joint, such as the elbow, dictates what angles are possible for the shoulder and vice versa.
+
+Expert roboticists manage high-dimensionality by representing the world in terms of {\em reference frames} (coordinate systems that are defined relative to particular robot part, such as the base or the head-mounted camera) and {\em quaternions} (a matrix that represents position and orientation in 3D space relative to a reference frame).
+Robot joint positions are typically specified relative to their own reference frame; for example, a robot's wrist, which is located at the end of its arm, can be commanded to rotate around the axis that is aligned with the robot's forearm (similar to the motion we use to screw in a light bulb).
+Calculating the absolute position of a robot's hand (for example, to determine whether it is near the door handle) requires a series of complex matrix calculations starting from the position of its base, through its shoulder, and all the joints, down to the hand.
+The opposite problem, calculating the joint positions required to get the hand to a particular position, is called {\em inverse kinematics} and is typically solved using heuristic approximation methods.
+More generally, techniques for {\em motion planning} calculate whole trajectories of how to move an arm from a starting location to an ending location, using inverse kinematics to compute trajectories for each of the joints in turn.
+
+Toolkits for doing automatic motion planning are on the verge of becoming mature, simplifying the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
+Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
+
+\section{Concurrency}
+
+% ROS nodes
+% multiple parts of robots (head, arms, base)
+% callbacks, coroutines, threads
+
+Robots are complex beings full of many moving parts.
+
+\section{Uncertainty}
+
+% World changes out from under you
+% Hardware fails
+% Incomplete world models
+% Bayesian methods, probabilities
 
 \section{ROS}
+
+Let's look at the state of the art in robot programming technology.
 
@@
 
-\section{Concurrency}
-
-ROS nodes
-multiple parts of robots (head, arms, base)
-callbacks, coroutines, threads
-
-\section{Uncertainty}
-
-World changes out from under you
-Hardware fails
-Incomplete world models
-Bayesian methods, probabilities
-
-\section{High-dimensionality}
-
-Robot arms have 7DOF
-grasps require 6D
-world is 3D
-quaternions, matrix manipulation
-specialized I/O devices
-
 \section{Speech interfaces}

### 2013-06-04 paper.tex
Notes
Uncertainty section

@@
 
+Another challenge that robot programmers face is specifying how the robot should behave in an uncertain world.
+The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around.
+People walk around, objects get lost, furniture gets moved.
+Moreover, limitations in current sensing technology make it difficult to build a perfectly accurate model of the world.
+For example, differences in lighting make the same object look very different to an RGB camera.
+Therefore a robot cannot necessarily trust that it what it perceives is reality, and it must be able to deal with the possibility that what it thinks is true is not actually true.
+
+What does this mean for robot programmers?
+Robot programs must have mechanisms in place to deal with uncertain truth.
+Rather than simplying instructing the robot to pick up an object, the program must specify what to do if the object is not where it thinks it is, or is not what it thinks it is.
+
+In addition, robot hardware itself often produces unpredictable results.
+Unlike software, mechanical parts can slip, burn out, or break.
+Giving a robot the same instruction multiple times may or may not result in the same result each time.
+
+In order to deal with these various types of uncertainty, professional programmers typically learn to use probabilistic models such as Bayesian networks or HMMs.
+However, these techniques require expert knowledge to use effectively and are out of the reach of casual programmers who just want to offload their work onto a robotic assistant.
+
+Robots that rely on having a model of the world in order to plan their actions must be able to deal with uncertainty both in the sensing of
+
 \section{ROS}
@@
 Before ROS, roboticists had to create the software to control their robots from scratch, each doing the same work to create device drivers and rebuild libraries for their specific robot's hardware.
+
+ROS runs on 50-100 robots
 

### 2013-06-17 paper.tex
More writing

@@
 
+TBD: talk about simplicity vs expressiveness tradeoff here?
+
 \section{High dimensionality}
@@
 Robots are complex beings full of many moving parts.
+For example, Willow Garage's robotic research platform, the PR-2, has two independently controlled arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
+Each of these parts must be controlled independently in order to orchestrate a desired robotic behavior.
+
+Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
+Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
+For many tasks this is a reasonable solution.
+However there is a subset of human tasks for which we need multiple parts moving simulateously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
+
+End user programming systems have tried to simplify the problem of concurrency.
+One approach used is visualizing the different components on a timeline.
+This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
+A different approach, taken by visual programming languages such as Scratch, enable programmers to define multiple blocks of code within the workspace, all of which run simultaneously and independently.
+
+A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
+For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
+As another example, a two-armed robot could have both arms linked in a mirror configuration so that a command to move one arm would cause the other to mirror its motion, thus enabling the robot to pick up symmetric objects with two hands.
+
+Even with higher-level primitives, there will still be a need for concurrent robot programming at a higher level.
+One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores.
+Supporting high level concurrency will introduce new problems, such as task prioritization (should it interrupt its task of bringing you a beer to clean up a wine spill on carpet?).
 
@@
 For example, differences in lighting make the same object look very different to an RGB camera.
-Therefore a robot cannot necessarily trust that it what it perceives is reality, and it must be able to deal with the possibility that what it thinks is true is not actually true.
-
-What does this mean for robot programmers?
-Robot programs must have mechanisms in place to deal with uncertain truth.
-Rather than simplying instructing the robot to pick up an object, the program must specify what to do if the object is not where it thinks it is, or is not what it thinks it is.
+Therefore a robot cannot necessarily trust what it perceives, and it must be able to deal with the possibility that what it thinks is true is not actually true.
 
@@
 
-In order to deal with these various types of uncertainty, professional programmers typically learn to use probabilistic models such as Bayesian networks or HMMs.
-However, these techniques require expert knowledge to use effectively and are out of the reach of casual programmers who just want to offload their work onto a robotic assistant.
+What does this mean for robot programmers?
+Robot programs must have mechanisms in place to deal with uncertain truth.
+Rather than simplying instructing the robot to pick up an object, the program must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
 
-Robots that rely on having a model of the world in order to plan their actions must be able to deal with uncertainty both in the sensing of
+In order to deal with these various types of uncertainty, professional programmers can learn to use probabilistic models such as Bayesian networks or HMMs, or they write code with lots of built-in error checking.
+However, casual programmers who just want a robot to perform a job in the home would find it burdensome to have to think through all the ways in which the robot could fail, and specify what the recovery strategy should be in each case.
 

### 2013-06-18 paper.tex
Added rviz section

@@
 \documentclass{article}
+\usepackage{graphicx}
+
 \begin{document}
@@
 
-TBD: talk about simplicity vs expressiveness tradeoff here?
+\section{The state of the art}
+
+% History of ROS
+% Invented by Willow Garage, new operating system for diverse robots
+% Current reach
+% Running on robots from PR2 to android-powered XXX
+% Provides very rich control substrate for programming the behavior of complex robots
+
+The Robot Operating System is one of the software advances that supports the rapid development of new robotics technology.
+First developed in 2007 at the Stanford Articifial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
+
+ROS provides a distributed publish/subscribe messaging platform for a collection of {\em nodes} to communicate with each other.
+Each node encapsulates a small amount of robot functionality, and interfaces with the rest of the ROS ecosystem to make that functionality available through a common interface.
+For example, ROS nodes exist to stream video from webcams, control the motors in robot arms, plan a path from one point to another, and recognize graspable objects in an image.
+
+Before ROS, roboticists had to create the software to control their robots from scratch, each doing the same work to create device drivers and rebuild libraries for their specific robot's hardware.
+Now, ROS has been ported to around 100 different models of robots worldwide\footnote{http://www.ros.org/wiki/Robots}, enabling roboticists to quickly bring a new robot model online.
+
+\begin{figure}[tbh]
+\scalebox{0.15}{\includegraphics{rviz.png}}
+\caption{RViz interface for controlling robots}
+\label{rviz}
+\end{figure}
+
+However, ROS provides a fairly low-level interface to robot control.
+Designed for roboticists, the RViz interface (Figure~\ref{rviz}) displays all the sensor feeds coming from the robot and lets one send low-level motion commands to the robot such as rolling forward or rotating the shoulder joint.
+In the figure, a robot is shown examining objects on a table in front of it.
+The green pattern represents the data from a depth sensor mounted on the robot's head, whereas the purple line represents the walls detected by a laser rangefinder mounted on the robot's base.
+The tricolored axes display the various coordinate frames in use, and the gray floor displays the map the robot uses for navigation.
+(TODO: display costmaps, TF axes)
+
+This interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly.
+It also illustrates some of the difficulties inherent in robot programming, which I will detail next.
 
@@
 
-\section{ROS}
-
-Let's look at the state of the art in robot programming technology.
-
-% History of ROS
-% Invented by Willow Garage, new operating system for diverse robots
-% Current reach
-% Running on robots from PR2 to android-powered XXX
-% Provides very rich control substrate for programming the behavior of complex robots
-
-The Robot Operating System is one of the software advances that supports the rapid development of new robotics technology.
-First developed in 2007 at the Stanford Articifial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
-
-ROS provides a distributed publish/subscribe messaging platform for a collection of {\em nodes} to communicate with each other.
-
-Before ROS, roboticists had to create the software to control their robots from scratch, each doing the same work to create device drivers and rebuild libraries for their specific robot's hardware.
-
-ROS runs on 50-100 robots
+\section{The way forward}
 

### 2013-06-18 paper.tex
More text

@@
 
+Most end user programming systems to date ignore the problem of uncertainty.
+Some can get away with this because they operate in a simulated world (such as Scratch).
+Others, which deal with uncertainty in the form of network requests timing or webservices being down, rely on the user to retry the operation as needed.
+A good user interface paradigm for handling uncertainty will be needed for end user programmed robot control software to be robust.
+
 \section{The way forward}

### 2013-07-02 paper.tex
First draft complete

@@
 
-\title {Robot programming for dummies}
+\title {Why robot programming is hard}
 \author {Tessa Lau \\
@@
 
-Actually, the problem is not that robots {\em can't} do what we want them to do.
-The problem is that it's too difficult to {\em tell} robots what we want them to do.
+Actually, the problem is not that robots {\em cannot} do what we want them to do.
+The problem is that it is too difficult to {\em tell} robots what we want them to do.
 The first company to crack the nut of end user robot programming will corner the market on household robots.
@@
 One potential use case for personal robotics will be in-home care and assistance for disabled adults.
-The Robots for Humanity project, a collaboration between Willow Garage and Georgia Tech, aims to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
+The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aims to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
 Its first prototypes have enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
@@
 
-The Robot Operating System is one of the software advances that supports the rapid development of new robotics technology.
+The Robot Operating System\cite{ros} is one of the software advances that supports the rapid development of new robotics technology.
 First developed in 2007 at the Stanford Articifial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
@@
 For many tasks this is a reasonable solution.
-However there is a subset of human tasks for which we need multiple parts moving simulateously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
+However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
 
@@
 
-Another challenge that robot programmers face is specifying how the robot should behave in an uncertain world.
+Another challenge that robot programmers face is specifying how a robot should behave in an uncertain world.
 The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around.
@@
 Unlike software, mechanical parts can slip, burn out, or break.
-Giving a robot the same instruction multiple times may or may not result in the same result each time.
+Giving a robot the same instruction multiple times may or may not result in the same outcome each time.
 
@@
 In order to deal with these various types of uncertainty, professional programmers can learn to use probabilistic models such as Bayesian networks or HMMs, or they write code with lots of built-in error checking.
-However, casual programmers who just want a robot to perform a job in the home would find it burdensome to have to think through all the ways in which the robot could fail, and specify what the recovery strategy should be in each case.
+However, casual programmers who just want a robot to perform a job in the home would find it burdensome to have to elaborate recovery strategies for each of the different failure conditions.
 
@@
 
-Why robotics is becoming important
-Statistics on size of market
-Number of research investments?
+Robot programming environments are beginning to address these challenges and produce useful systems that can be controlled by untrained users.
+The key tradeoff is simplicity versus expressiveness: in order to make systems that are more tractable for novices, they need to simplify key aspects of the robot's behavior, thereby sacrificing its ability to perform a full range of behavior.
 
-Main points:
-	EUP is what will enable robotics to spawn new industries
-	by enabling businesses to customize robot behavior for their own needs
+This past generation of commercial home robots illustrates one extreme end of the simplicity/expressiveness tradeoff.
+iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''.
 
-	Yet programming robots is difficult
-	State of the art today is demonstrational, which works well for limited, repetitive tasks
-	Or using highly technical tools to specify fine-grained robot behavior
-	Behavior specification requires much deep technical knowledge
-	Not only the traditionally hard topics in EUP such as conditionals, loops, variables
-	But also advanced topics like concurrency, uncertainty, high-dimensionality
+The next generation of commercial robots in this space will likely look like Rethink Robots' Baxter system\footnote{http://www.rethinkrobotics.com/index.php/products/baxter/}.
+Baxter is a two-armed manipulator than can be programmed by demonstration to automate factory work simply by guiding its arms to the appropriate position and tapping icons on the robot's ``face''.
+This style of programming makes it possible for people to specify the robot's behavior directly on the robot, without having to deal with any abstractions.
+In addition, Baxter performs its work in a human-safe way, immediately ceasing motion if it detects that a human has moved into its path, enabling it to work side by side with factory workers unlike previous generations of industrial robots.
 
-The challenge for robotics is to bring robots to market that are consumer-friendly and can operate in human environments.
+For the future, our team at Willow Garage is exploring different tradeoff points in the space of simplicity versus expressiveness.
+One research direction we have been exploring is to look at specific verticals and design a customized set of robot programming primitives specifically for that domain.
+Each of these primitives can be hand-crafted in the lab by expert roboticists to embody a rich amount of automation, parallelism, and support for uncertainty.
+The end user programming problem is then a simpler matter of sequencing these higher-level primitives into useful, customizable programs for their environment.
+
+For example, our team has been doing ethnographic investigations of employees in continuing care retirement centers to discover how robotic helpers could enable them to spend more time with residents.
+We have discovered that a robotic assistant who could autonomously fetch and deliver supplies to arbitrary locations within the facility could be beneficial to their workflow.
+Based on those findings, we have defined a set of primitives that include picking up and dropping off supply bins, navigating autonomously to pre-defined locations, and prompting for user input via a touchscreen on the robot's ``face''.
+End user programmers will be able to sequence these primitives themselves to design custom behaviors for a robotic delivery system, while expert roboticists can focus on making individual primitives robust using the full slate of advanced programming techniques at their disposal.
+
+Over time, as we discover and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior.
+And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.
+
+% Why robotics is becoming important
+% Statistics on size of market
+% Number of research investments?
+% 
+% Main points:
+% 	EUP is what will enable robotics to spawn new industries
+% 	by enabling businesses to customize robot behavior for their own needs
+% 
+% 	Yet programming robots is difficult
+% 	State of the art today is demonstrational, which works well for limited, repetitive tasks
+% 	Or using highly technical tools to specify fine-grained robot behavior
+% 	Behavior specification requires much deep technical knowledge
+% 	Not only the traditionally hard topics in EUP such as conditionals, loops, variables
+% 	But also advanced topics like concurrency, uncertainty, high-dimensionality
+% 
+% The challenge for robotics is to bring robots to market that are consumer-friendly and can operate in human environments.
 For limited tasks, we are seeing robots that can perform one single task very well without getting in the way of the humans in the household.
@@
 Undoubtedly their success is due to their extremely simple interface: the Roomba has a single large button labeled ``Clean''.
-
-read this: http://workshop.iroboticist.com/why-robotics/
-
-\section{Related work}
-
-Roomba, Neato in consumer use
-	http://www.neatorobotics.com/
-	http://store.irobot.com/home/index.jsp
-Baxter human-safe 2-armed manipulator
-	runs ROS
-	http://www.ros.org/news/2012/09/rethink-ros.html
-	http://www.rethinkrobotics.com/index.php/products/baxter/
-Industrial robots:
-	KUKA robot arm http://en.wikipedia.org/wiki/KUKA
-Toyota's human support robot:
-	runs ROS
-	http://www.gizmag.com/toyota-human-support-robot/24246/
-Telepresence robots:
-	Beam
-	Anybot QB
-	Double
-
-\section{Speech interfaces}
-
-Impedance mismatch of speech interfaces
+% 
+% 
+% read this: http://workshop.iroboticist.com/why-robotics/
+% 
+% \section{Related work}
+% 
+% Roomba, Neato in consumer use
+% 	http://www.neatorobotics.com/
+% 	http://store.irobot.com/home/index.jsp
+% Baxter human-safe 2-armed manipulator
+% 	runs ROS
+% 	http://www.ros.org/news/2012/09/rethink-ros.html
+% 	http://www.rethinkrobotics.com/index.php/products/baxter/
+% Industrial robots:
+% 	KUKA robot arm http://en.wikipedia.org/wiki/KUKA
+% Toyota's human support robot:
+% 	runs ROS
+% 	http://www.gizmag.com/toyota-human-support-robot/24246/
+% Telepresence robots:
+% 	Beam
+% 	Anybot QB
+% 	Double
+% 
+% 
+% \section{Speech interfaces}
+% 
+% Impedance mismatch of speech interfaces
 

### 2013-07-18 paper.tex
Checkpoint.
Folded in comments from Kaijen and Andreas. Starting to make the message come out more strongly in the intro and case study.

@@
-\documentclass{article}
+\documentclass[11pt,twocolumn]{article}
 \usepackage{graphicx}
@@
 Far better to work in the world of software, I figured, where the systems I built could make a real difference in people's lives.
-Software robots would automate repetitive tasks in computer use and free people from mundane work.
-End user programming would enable millions of ordinary computer users to customize and adapt systems to their own needs.
+Software could automate repetitive tasks in computer use and free people from mundane work.
+My chosen field, end user programming, had the potential to enable millions of ordinary computer users to customize and adapt systems to their own needs.
 
-But that world is changing.
-Computing has moved off the desktop and into the smartphones, smart environments, and the world aronud us.
-Innovation is happening in the real world, using computing to effect physical change in our environment.
+But the world is changing.
+Computing has moved off the desktop and into the smartphones, smart environments, and the world around us.
+Innovation is happening in the physical world, using computing to effect material change in our environment.
 The past decades have seen industrial robots increase manufacturing efficiency on the factory floor.
@@
 Yet the holy grail of personal robotics remains elusive.
-When will we have our own Rosie the Robot (from the Jetsons cartoon) to take over housekeeping chores?
+When will we have our own Rosie the Robot to take over housekeeping chores?
 When will disabled adults have robot helpers to assist them in activities of daily living?
@@
 
-Actually, the problem is not that robots {\em cannot} do what we want them to do.
+What I have found is that the problem is not that robots {\em cannot} do what we want them to do.
 The problem is that it is too difficult to {\em tell} robots what we want them to do.
-The first company to crack the nut of end user robot programming will corner the market on household robots.
-Human-robot interaction is the missing link for widespread personal robotics.
+Robot control can be roughly categorized into three distinct phases: direct demonstration, tele-operation (or teleop for short), and higher-level programming.
+
+In demonstrational interfaces, a human operator moves a robot through a series of positions to demonstrate the motion it should perform.
+The robot then continues to perform this motion unattended.
+This style of programming is common for industrial robots, which can be programmed to move items around on conveyor belts or do precise assembly tasks.
+Demonstrational interfaces are typically limited to repeating only what was previously demonstrated, in the worst case performing the exact same motion down to the millimeter regardless of whether the part is in the same place as it was last time.
+
+Teleop consists of synchronously controlling a robot using software and hardware interfaces from a computer which may or may not be co-located with the robot.
+For example, a PS3 joystick controller could be used to drive a robot around a building, and activate certain pre-programmed motions such as turning its head or extending its arms.
+Remote teleop consists of controlling the robot over a distance, perhaps over the internet, using GUIs that allow the human operator to move various parts of the robot.
+
+Both demonstration and teleop suffer the limitation that a human operator must be immediately in the loop while the robot is operating.
+To make robots useful in the home, they will need to function {\em autonomously}, which leads to the notion of robot programming: giving high-level directives to robots so that they can function on their own without direct human guidance.
+A full programming solution will likely involve a blend of different techniques including demonstration, teleop, and autonomous subroutines.
+For example, a task of setting the dinner table may include a high-level program of fetching silverware from the kitchen and placing it on the dining table, a demonstration of how to open this particular style of cabinet, and teleoperation when something goes awry (such as the silverware drawer being empty).
+
+For personal robotics to truly succeed, robot programming will need to be simple enough that non-experts can do it.
 
@@
 
-One potential use case for personal robotics will be in-home care and assistance for disabled adults.
+One important use case for personal robotics will be in-home care and assistance for disabled adults.
 The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aims to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
 Its first prototypes have enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
-Limited to the ability to move a single finger, Henry has been able to instruct Willow Garage's PR2 robots to open cabinets, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+Through physical therapy, Henry has regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
+Using a teleop interface custom designed for him, Henry has been able to instruct Willow Garage's PR2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+
+% TODO: conflating robot programming with teleop
 
 Yet our experience with Robots for Humanity underscores how difficult it is for ordinary mortals to instruct robots what to do.
-Even using state-of-the-art robot visualization software developed at Willow Garage, simple tasks such as opening a cabinet door could take Henry upwards of 15 minutes to accomplish.
-While Henry has only limited interaction capabilities due to motor impairment, he also has incredible motivation.
-Ordinary consumers, on the other hand, while fully able-bodied, may not share the same perseverance in programming robots.
+Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door could take Henry upwards of 15 minutes to accomplish.
+While Henry has only limited interaction capabilities due to motor impairment, he is also strongly motivated.
+Ordinary able-bodied consumers, on the other hand, would quickly lose patience.
+If basic robot control via teleoperation is so time-consuming, what hope do we have that people will ever get around to the general task of robot programming?
+
+Granted, designing programming interfaces for motor impaired users is like playing the piano with both hands tied behind your back.
+Yet while Henry's case highlights the challenges of creating usable interfaces for assistive robotics, it also showcases the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
 
@@
 The Robot Operating System\cite{ros} is one of the software advances that supports the rapid development of new robotics technology.
-First developed in 2007 at the Stanford Articifial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
+First developed in 2007 at the Stanford Artificial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
 
@@
 This interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly.
-It also illustrates some of the difficulties inherent in robot programming, which I will detail next.
+It also illustrates some of the difficulties inherent in robot programming, described below.
 
@@
 High dimensionality is even more important in specifying how to grasp and place objects in the world.
+In many cases, the robot itself might occlude the object it is trying to manipulate.
+For example, if a robot uses its arm to pick up an object on a table in front of it, the arm might get in the way of a sensor mounted on its head, preventing the robot from seeing the exact thing it is trying to grasp!
+One solution developed by roboticists has been to include additional cameras on different parts of the robot, such as a forearm-mounted camera, to get a better view on the object being picked up.
+However, making sense of multiple translated and rotated video streams can be confusing for those of us accustomed to seeing only from our own head-mounted eye sensors!
+
+High dimensionality also makes it difficult to specify to how pick up objects using a robotic gripper.
 A solid object has six degrees of freedom: three degrees to specify its position, and another three to specify its rotation.
@@
 These interfaces are cumbersome because the joints are interdependent: setting the angle of one joint, such as the elbow, dictates what angles are possible for the shoulder and vice versa.
+There is a tradeoff between exposing these joint limitations to the programmer, which allows precise yet complex control, versus insulating a programmer from those considerations, which may lead to unpredictable behavior that is difficult to debug.
 
-Expert roboticists manage high-dimensionality by representing the world in terms of {\em reference frames} (coordinate systems that are defined relative to particular robot part, such as the base or the head-mounted camera) and {\em quaternions} (a matrix that represents position and orientation in 3D space relative to a reference frame).
+Expert roboticists manage high-dimensionality by representing the world in terms of {\em reference frames} (coordinate systems that are defined relative to particular robot part, such as the base or the head-mounted camera) and use matrix algebra to convert between different frames of reference.
 Robot joint positions are typically specified relative to their own reference frame; for example, a robot's wrist, which is located at the end of its arm, can be commanded to rotate around the axis that is aligned with the robot's forearm (similar to the motion we use to screw in a light bulb).
 Calculating the absolute position of a robot's hand (for example, to determine whether it is near the door handle) requires a series of complex matrix calculations starting from the position of its base, through its shoulder, and all the joints, down to the hand.
-The opposite problem, calculating the joint positions required to get the hand to a particular position, is called {\em inverse kinematics} and is typically solved using heuristic approximation methods.
-More generally, techniques for {\em motion planning} calculate whole trajectories of how to move an arm from a starting location to an ending location, using inverse kinematics to compute trajectories for each of the joints in turn.
+The opposite problem, calculating the joint positions required to get the hand to a particular position, is called {\em inverse kinematics} and is typically solved using either analytical equations or heuristic approximation methods.
+
+% TODO: introduce DOF acronym earlier
 
@@
 Unlike software, mechanical parts can slip, burn out, or break.
+They take time to actuate and do not come to a stop immediately.
 Giving a robot the same instruction multiple times may or may not result in the same outcome each time.
@@
 What does this mean for robot programmers?
-Robot programs must have mechanisms in place to deal with uncertain truth.
+Robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware.
+Robot control loops usually sense the physical hardware during control to determine whether a motion was carried out successfully.
+% TODO
 Rather than simplying instructing the robot to pick up an object, the program must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.

### 2013-07-18 paper.tex
Another checkpoint. Working through the challenges sections.

@@
 A full programming solution will likely involve a blend of different techniques including demonstration, teleop, and autonomous subroutines.
-For example, a task of setting the dinner table may include a high-level program of fetching silverware from the kitchen and placing it on the dining table, a demonstration of how to open this particular style of cabinet, and teleoperation when something goes awry (such as the silverware drawer being empty).
+For example, a task of setting the dinner table may include a high-level program of fetching silverware from the kitchen and placing it on the dining table, a demonstration of how to open this particular silverware drawer, and teleoperation when something goes awry (such as the silverware drawer being empty).
 
@@
 Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door could take Henry upwards of 15 minutes to accomplish.
-While Henry has only limited interaction capabilities due to motor impairment, he is also strongly motivated.
-Ordinary able-bodied consumers, on the other hand, would quickly lose patience.
-If basic robot control via teleoperation is so time-consuming, what hope do we have that people will ever get around to the general task of robot programming?
+While Henry has only limited interaction capabilities, he is also strongly motivated because he has no alternatives.
+Able-bodied consumers, on the other hand, would quickly lose patience.
+If even basic robot control via teleoperation is so time-consuming, what hope do we have that people will ever get around to the general task of robot programming?
 
-Granted, designing programming interfaces for motor impaired users is like playing the piano with both hands tied behind your back.
-Yet while Henry's case highlights the challenges of creating usable interfaces for assistive robotics, it also showcases the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
+Granted, designing teleoperation interfaces for motor-impaired users is deliberately trying to solve the hardest problem first.
+Yet while Henry's case highlights the challenges of creating usable assistive robotics interfaces, it also illustrates the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
 
-Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
+With this goal in mind, autonomous programmable robots are more than a consumer toy, but a technology that could changes people's lives for the better.
+Why, then, does robot programming for the masses remain such a difficult challenge?
+What makes robots so difficult to control?
+
+The explanation begins with a look at the state of the art in robot software today.
+
+% Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
 Why is end user robot programming so challenging?
@@
 Each node encapsulates a small amount of robot functionality, and interfaces with the rest of the ROS ecosystem to make that functionality available through a common interface.
-For example, ROS nodes exist to stream video from webcams, control the motors in robot arms, plan a path from one point to another, and recognize graspable objects in an image.
+For example, ROS nodes exist to stream sensor data from webcams and laser rangefinders, control the motors in robot arms, plan a path from one point to another, and recognize graspable objects in an image.
 
@@
 
-\begin{figure}[tbh]
-\scalebox{0.15}{\includegraphics{rviz.png}}
+\begin{figure*}[tbh]
+\center\scalebox{0.20}{\includegraphics{rviz.png}}
 \caption{RViz interface for controlling robots}
 \label{rviz}
-\end{figure}
+\end{figure*}
 
@@
 This interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly.
-It also illustrates some of the difficulties inherent in robot programming, described below.
+It also illustrates some of the difficulties inherent in robot programming.
 
-\section{High dimensionality}
+\section{A 3D world}
 
@@
 
-Unlike traditional computer interfaces which are restricted to two dimensions, robot control requires manipulating objects in a three-dimensional world.
+Unlike traditional computer interfaces which are limited to two dimensions, robot control requires manipulating objects in a three-dimensional world.
 Visualizing a 3D world on existing 2D displays requires the extra dimension to be projected down into two dimensions, giving rise to the popular ``first person shooter'' style of interface.
 These interfaces display the world as perceived from a virtual camera, where objects can be occluded by other objects.
-Additional camera controls are needed to pan/zoom around the space to overcome occlusion, and these controls can be nonintuitive to non-expert computer users.
+Additional camera controls are needed to pan/zoom around the space to overcome occlusion.
+These controls are often non-intuitive, even to seasoned computer users.
 
-High dimensionality is even more important in specifying how to grasp and place objects in the world.
-In many cases, the robot itself might occlude the object it is trying to manipulate.
-For example, if a robot uses its arm to pick up an object on a table in front of it, the arm might get in the way of a sensor mounted on its head, preventing the robot from seeing the exact thing it is trying to grasp!
-One solution developed by roboticists has been to include additional cameras on different parts of the robot, such as a forearm-mounted camera, to get a better view on the object being picked up.
-However, making sense of multiple translated and rotated video streams can be confusing for those of us accustomed to seeing only from our own head-mounted eye sensors!
+Occlusion presents a challenge because robots must perceive an object in order to manipulate it.
+Often times, the robot itself might occlude the object it is trying to manipulate!
+For example, consider a humanoid robot with head-mounted vision sensors using an arm to pick up an object from the table in front of it.
+While moving the arm into position to grasp the object, the arm will likely come between the head and the object, obscuring the robot's vision.
+
+One solution has been to mount additional cameras on different parts of the robot, such as on the robot's forearm or wrist, to get a better view on the object being picked up.
+However, making sense of multiple translated and rotated video streams can be confusing for those of us accustomed to perceiving the world only through head-mounted eye sensors!
+
+Typical input devices such as mice and touchscreens are also only good for interacting in two dimensions.
+For example, tapping on a point in a camera image does not uniquely identify a point in 3D space, but a line of possible points.
+When telling a robot to put something down, how can it know which of those points was meant?
+Current interfaces, such as the RViz display shown in Figure~\ref{rviz}, require complex camera manipulation and precise 3D positioning in order to uniquely specify the desired location.
+
+\section{Arm complexity}
+
+A different aspect of high-dimensionality comes from the design of robot arms.
+A solid object in 3D space has six degrees of freedom (DOF): three dimensions to specify its position, and three more to specify its rotation.
+If a robot arm had only 6-DOF and was trying to execute a specific grasp of a particular object, such as picking up soda can from the side, there would be only a single arm configuration that would allow the gripper to pick up the can.
+In contrast, human arms and advanced robot arms have 7-DOF, which allows them one extra dimension of flexibility in achieving those grasps.
+To visualize this, imagine grasping a soda can on the table in front of you and then flapping your elbow up and down while keeping the can stationary.
+That extra degree of freedom enables your arm to take on all those different shapes while still achieving the desired grasp.
+
+Interfaces for robot arm teleoperation typically involve setting the position or angle of each joint in the arm individually.
+For example, a tablet-style interface might have fourteen buttons, one pair used to increase or decrease the angle of each joint in the arm.
+That additional degree of freedom makes it easier to identify a valid arm configuration for executing a grasp, since only one of many valid configurations must be found, not the unique one.
+
+For teleoperators who are manually positioning a robot arm must only find one way to position the various joints
+
+It may also make it easier for teleoperators to find a valid arm configuration to reach a desired grasp.
+Yet those seven degrees of freedom also represent seven variables to control when using a direct-manipulation teleop interface
 
@@
 The opposite problem, calculating the joint positions required to get the hand to a particular position, is called {\em inverse kinematics} and is typically solved using either analytical equations or heuristic approximation methods.
-
-% TODO: introduce DOF acronym earlier
 

### 2013-07-18 paper.tex
Done for the day. Still need to revisit the Concurrency section, but I've made a full pass over the rest.

@@
 A solid object in 3D space has six degrees of freedom (DOF): three dimensions to specify its position, and three more to specify its rotation.
-If a robot arm had only 6-DOF and was trying to execute a specific grasp of a particular object, such as picking up soda can from the side, there would be only a single arm configuration that would allow the gripper to pick up the can.
+If a robot arm had only 6-DOF and was trying to execute a specific grasp of a particular object, such as picking up soda can from the side, there would be only a single arm configuration that would allow the gripper (a robotic hand) to pick up the can.
 In contrast, human arms and advanced robot arms have 7-DOF, which allows them one extra dimension of flexibility in achieving those grasps.
@@
 That additional degree of freedom makes it easier to identify a valid arm configuration for executing a grasp, since only one of many valid configurations must be found, not the unique one.
+However controlling each individual joint position by clicking on buttons is tedious.
+Additionally, since the allowable joint positions are not completely independent, finding a correct solution is often done through trial and error.
 
-For teleoperators who are manually positioning a robot arm must only find one way to position the various joints
+%There is a tradeoff between exposing these joint limitations to the programmer, which allows precise yet complex control, versus insulating a programmer from those considerations, which may lead to unpredictable behavior that is difficult to debug.
 
-It may also make it easier for teleoperators to find a valid arm configuration to reach a desired grasp.
-Yet those seven degrees of freedom also represent seven variables to control when using a direct-manipulation teleop interface
+Yet positioning an arm through direct manipulation or teleop is relatively easy compared to doing it programmatically.
+In order to program an arm, one must first specify {\em where} it should go.
+Expert roboticists represent the world in terms of {\em reference frames} (coordinate systems that are defined relative to a fixed position).
+For example, one reference frame could be relative to the building, with the front door being at coordinate (0, 0).
+Other reference frames are defined relative to stationary parts of the robot, such as its base (if it is on wheels), the head-mounted camera, or the right elbow joint.
 
-High dimensionality also makes it difficult to specify to how pick up objects using a robotic gripper.
-A solid object has six degrees of freedom: three degrees to specify its position, and another three to specify its rotation.
-The human arm, and the most advanced robotic arms, have seven degrees of freedom.
-Many robot arm control interfaces ask users to independently specify joint angles for each of those seven degrees of freedom.
-These interfaces are cumbersome because the joints are interdependent: setting the angle of one joint, such as the elbow, dictates what angles are possible for the shoulder and vice versa.
-There is a tradeoff between exposing these joint limitations to the programmer, which allows precise yet complex control, versus insulating a programmer from those considerations, which may lead to unpredictable behavior that is difficult to debug.
+Sending an arm to a specific position is useful for robot tasks such as, for example, looking at what it is holding by moving its hand to half a meter in front of its face.
+This half-meter is most reliably specified in the reference frame relative to the robot's head.
+Given this goal position, the programmer must find a series of arm configurations that moves the arm to the desired place, while avoiding contact with obstacles in the environment.
 
-Expert roboticists manage high-dimensionality by representing the world in terms of {\em reference frames} (coordinate systems that are defined relative to particular robot part, such as the base or the head-mounted camera) and use matrix algebra to convert between different frames of reference.
+% such as the base or the head-mounted camera) and use matrix algebra to convert between different frames of reference.
 Robot joint positions are typically specified relative to their own reference frame; for example, a robot's wrist, which is located at the end of its arm, can be commanded to rotate around the axis that is aligned with the robot's forearm (similar to the motion we use to screw in a light bulb).
@@
 
-Toolkits for doing automatic motion planning are on the verge of becoming mature, simplifying the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
+Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
 Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
@@
 Robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware.
-Robot control loops usually sense the physical hardware during control to determine whether a motion was carried out successfully.
-% TODO
-Rather than simplying instructing the robot to pick up an object, the program must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
+Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
+Unlike with demonstration and teleop, where the operator can see if the robot carried out the desired motion and take appropriate action, a robot programmer must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
+For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
+Should it simply try again?
+Move a little bit to get a better view and then try again?
+Ask a human?
 
-In order to deal with these various types of uncertainty, professional programmers can learn to use probabilistic models such as Bayesian networks or HMMs, or they write code with lots of built-in error checking.
-However, casual programmers who just want a robot to perform a job in the home would find it burdensome to have to elaborate recovery strategies for each of the different failure conditions.
-
-Most end user programming systems to date ignore the problem of uncertainty.
-Some can get away with this because they operate in a simulated world (such as Scratch).
-Others, which deal with uncertainty in the form of network requests timing or webservices being down, rely on the user to retry the operation as needed.
-A good user interface paradigm for handling uncertainty will be needed for end user programmed robot control software to be robust.
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and learn to craft fault-tolerant code with frequent error checking, which seems like an heavy burden for casual programmers.
+Most end user programming systems to date simply ignore the problem of uncertainty.
+For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.
 
@@
 
-This past generation of commercial home robots illustrates one extreme end of the simplicity/expressiveness tradeoff.
+This past generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff.
 iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''.
+While the robot will eventually clean your floors, you lose control over how exactly it does so, as it performs what seems like a random walk through your home.
 
-The next generation of commercial robots in this space will likely look like Rethink Robots' Baxter system\footnote{http://www.rethinkrobotics.com/index.php/products/baxter/}.
+%The next generation of commercial robots in this space will likely look like Rethink Robots' Baxter system\footnote{http://www.rethinkrobotics.com/index.php/products/baxter/}.
 Baxter is a two-armed manipulator than can be programmed by demonstration to automate factory work simply by guiding its arms to the appropriate position and tapping icons on the robot's ``face''.
-This style of programming makes it possible for people to specify the robot's behavior directly on the robot, without having to deal with any abstractions.
+Unlike past demonstration-based robots which mindlessly repeated the same motions, Baxter has an intelligent motion planner which can immediately cease moving the arms if it detects something in the way, such as a human walking into its workspace.
+
+%This style of programming makes it possible for people to specify the robot's behavior directly on the robot, without having to deal with any abstractions.
 In addition, Baxter performs its work in a human-safe way, immediately ceasing motion if it detects that a human has moved into its path, enabling it to work side by side with factory workers unlike previous generations of industrial robots.
 
-For the future, our team at Willow Garage is exploring different tradeoff points in the space of simplicity versus expressiveness.
-One research direction we have been exploring is to look at specific verticals and design a customized set of robot programming primitives specifically for that domain.
-Each of these primitives can be hand-crafted in the lab by expert roboticists to embody a rich amount of automation, parallelism, and support for uncertainty.
-The end user programming problem is then a simpler matter of sequencing these higher-level primitives into useful, customizable programs for their environment.
+Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles), and recognizing and picking up discrete objects on a flat surface.
+Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own, but it also reduces expressivity by limiting the ways in which those tasks can be performed to those that are programmed in to the automation.
 
-For example, our team has been doing ethnographic investigations of employees in continuing care retirement centers to discover how robotic helpers could enable them to spend more time with residents.
-We have discovered that a robotic assistant who could autonomously fetch and deliver supplies to arbitrary locations within the facility could be beneficial to their workflow.
-Based on those findings, we have defined a set of primitives that include picking up and dropping off supply bins, navigating autonomously to pre-defined locations, and prompting for user input via a touchscreen on the robot's ``face''.
-End user programmers will be able to sequence these primitives themselves to design custom behaviors for a robotic delivery system, while expert roboticists can focus on making individual primitives robust using the full slate of advanced programming techniques at their disposal.
+Our approach to designing end-user-programmable interfaces is to craft these autonomous components into a good set of {\em primitives} in an end user programming language, such that a non-expert would not have to ``look inside the box'' to understand how each one works.
+Roboticists can then implement these primitives using advanced techniques to solve the difficult problems encountered during robot programming.
+For example, a good autonomous navigation component could incorporate a set of recovery behaviors to try on navigation failure, such as turning around, waiting for people/obstacles to clear out of its path, and backing up and retrying.
+It could also bake in some concurrent strategies, such as turning the head to look in the direction the robot is going.
 
-Over time, as we discover and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior.
+%Encapsulating all this functionality into a single component with a well-defined interface makes it easier for end user robot programmers by giving them more reliable, easier to understand components they can assemble into programs.
+
+We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops that enables casual users to control our PR2 robots through a web interface (Figure~\ref{contops}).
+This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
+
+This interface lies somewhere between teleoperation and programming.
+It does not require an operator to exert direct control over the robot, but the commands that are given are executed immediately by the robot rather than scheduled to occur later in time.
+We are also exploring interfaces for sequencing these primitives into scheduled programs, in order to validate that the set of primitives we are developing is complete enough to perform useful tasks.
+
+Another key component of our approach, which we call {\em shared autonomy}, is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future.
+In our system designs, the teleoperator does not necessarily have to be the person who wrote the program.
+Instead, we are modeling our system after telephone call centers where a team of trained professionals handle incidents as they arise and get the robots back on track to fulfill their programmed goals.
+As a stepping-stone towards full autonomy, this approach enables us to start testing end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.
+
+Over time, as we refine existing autonomous primitives and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior.
 And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.

### 2013-07-19 paper.tex
Concurrency section rewrite

@@
-\documentclass[11pt,twocolumn]{article}
+\documentclass[10pt,twocolumn]{article}
 \usepackage{graphicx}
@@
 Through physical therapy, Henry has regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
-Using a teleop interface custom designed for him, Henry has been able to instruct Willow Garage's PR2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+Using a teleop interface custom designed for him, Henry has been able to instruct Willow Garage's PR-2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
 
@@
 
-\section{Arm complexity}
+\section{7-DOF arms}
 
@@
 Robots are complex beings full of many moving parts.
-For example, Willow Garage's robotic research platform, the PR-2, has two independently controlled arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
+For example, Willow Garage's robotic research platform, the PR-2, has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
 Each of these parts must be controlled independently in order to orchestrate a desired robotic behavior.
@@
 This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
-A different approach, taken by visual programming languages such as Scratch, enable programmers to define multiple blocks of code within the workspace, all of which run simultaneously and independently.
+A second paradigm, taken by visual programming languages such as Scratch, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
+More research is needed to discover what paradigm works best for controlling robot parts concurrently.
 
-A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
+% A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
 For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
@@
 
-Even with higher-level primitives, there will still be a need for concurrent robot programming at a higher level.
+Even if individual robot parts can be programmed to act in parallel, there will still be a need for reasoning about concurrent robot behavior at a higher level.
 One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores.
@@
 
-We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops that enables casual users to control our PR2 robots through a web interface (Figure~\ref{contops}).
-This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
+We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops that enables casual users to control our PR-2 robots through a web interface (Figure~\ref{contops}).
+This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR-2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
 

### 2013-07-19 paper.tex
Version 2 shipped.

@@
 \center\scalebox{0.20}{\includegraphics{rviz.png}}
-\caption{RViz interface for controlling robots}
+\caption{RViz interface for robot visualization}
 \label{rviz}
@@
 
-\section{Concurrency}
-
-% ROS nodes
-% multiple parts of robots (head, arms, base)
-% callbacks, coroutines, threads
-
-Robots are complex beings full of many moving parts.
-For example, Willow Garage's robotic research platform, the PR-2, has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
-Each of these parts must be controlled independently in order to orchestrate a desired robotic behavior.
-
-Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
-Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
-For many tasks this is a reasonable solution.
-However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
-
-End user programming systems have tried to simplify the problem of concurrency.
-One approach used is visualizing the different components on a timeline.
-This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
-A second paradigm, taken by visual programming languages such as Scratch, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
-More research is needed to discover what paradigm works best for controlling robot parts concurrently.
-
-% A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
-For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
-As another example, a two-armed robot could have both arms linked in a mirror configuration so that a command to move one arm would cause the other to mirror its motion, thus enabling the robot to pick up symmetric objects with two hands.
-
-Even if individual robot parts can be programmed to act in parallel, there will still be a need for reasoning about concurrent robot behavior at a higher level.
-One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores.
-Supporting high level concurrency will introduce new problems, such as task prioritization (should it interrupt its task of bringing you a beer to clean up a wine spill on carpet?).
-
 \section{Uncertainty}
@@
 
+\section{Concurrency}
+
+% ROS nodes
+% multiple parts of robots (head, arms, base)
+% callbacks, coroutines, threads
+
+Robots are complex beings full of many moving parts.
+For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
+Robots are also equipped with a slew of sensors including cameras, depth sensors, laser rangefinders, touch sensors, and feedback about its electrical and motor systems.
+Sensors and actuators must be operated together in coordination to produce a desired robotic behavior.
+
+Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
+Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
+For many tasks this is a reasonable solution.
+However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
+
+End user programming systems have tried to simplify the problem of concurrency.
+Some systems visualize the different components on a timeline.
+This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
+A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
+More research is needed to discover what paradigm works best for controlling robot parts concurrently.
+
+% A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
+For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
+As another example, a two-armed robot could have both arms linked in a mirror configuration so that a command to move one arm would cause the other to mirror its motion, thus enabling the robot to pick up symmetric objects with two hands.
+
+Even if individual robot parts can be programmed to act in parallel, there will still be a need for reasoning about concurrent robot behavior at a higher level.
+One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores.
+Supporting high level concurrency will introduce new problems, such as task prioritization (is the current task of bringing you a beer more important than cleaning up the wine spill it just detected?).
+
 \section{The way forward}
 
-Robot programming environments are beginning to address these challenges and produce useful systems that can be controlled by untrained users.
+Robot programming environments are beginning to address all of these challenges and produce useful systems that can be controlled by untrained users.
 The key tradeoff is simplicity versus expressiveness: in order to make systems that are more tractable for novices, they need to simplify key aspects of the robot's behavior, thereby sacrificing its ability to perform a full range of behavior.
@@
 
-Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles), and recognizing and picking up discrete objects on a flat surface.
-Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own, but it also reduces expressivity by limiting the ways in which those tasks can be performed to those that are programmed in to the automation.
+Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles)\footnote{MoveIt!, http://moveit.ros.org}, and recognizing and picking up discrete objects on a flat surface.
+Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own.
 
@@
 
-We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops that enables casual users to control our PR-2 robots through a web interface (Figure~\ref{contops}).
+\begin{figure*}[tbh]
+\center\scalebox{0.40}{\includegraphics{contops.png}}
+\caption{Continuous Ops interface for picking up objects on a table}
+\label{contops}
+\end{figure*}
+
+We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enables casual users to control our PR-2 robots through a web interface (Figure~\ref{contops}).
 This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR-2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.

### 2013-07-26 paper.tex
Notes from Matei's feedback

@@
 \maketitle
+
+% Feedback from Matei:
+
+% Make assumptions more explicit in intro
+% Robot use == robot programming
+% Differentiate end user programming from roboticist programming
+% Teach/instruct/tell == programming
+
+% Non-quantitative Henry statistics
+
+% Cartesian robot control is a solved problem
 
@@
 
-% TODO: conflating robot programming with teleop
-
 Yet our experience with Robots for Humanity underscores how difficult it is for ordinary mortals to instruct robots what to do.
-Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door could take Henry upwards of 15 minutes to accomplish.
+Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door could be very difficult for Henry to accomplish.
 While Henry has only limited interaction capabilities, he is also strongly motivated because he has no alternatives.

### 2013-11-20 paper.tex
Final version as submitted

@@
 \author {Tessa Lau \\
-Willow Garage \\
-Menlo Park, CA USA \\
-{\tt tlau@willowgarage.com}
+Affiliation TBD \\
+Buenos Aires, Argentina \\
+{\tt tlau@tlau.org}
 }
@@
 The problem is that it is too difficult to {\em tell} robots what we want them to do.
-Robot control can be roughly categorized into three distinct phases: direct demonstration, tele-operation (or teleop for short), and higher-level programming.
+Even basic robot use entails some form of programming, because autonomous algorithms are not yet reliable and robust enough to complete useful tasks.
+Except in a few narrow domains where automation has become robust and practical, the mere act of telling a robot what to do involves a complex set of skills.
+
+Robot control can be roughly categorized into three distinct phases: direct demonstration, tele-operation (or teleop for short), and end user programming.
 
@@
 This style of programming is common for industrial robots, which can be programmed to move items around on conveyor belts or do precise assembly tasks.
-Demonstrational interfaces are typically limited to repeating only what was previously demonstrated, in the worst case performing the exact same motion down to the millimeter regardless of whether the part is in the same place as it was last time.
+Demonstrational interfaces are typically limited to repeating only what was previously demonstrated, performing the exact same motion down to the millimeter.
+While this is useful in controlled environments such as factories, typical home environments are much more fluid and changing.
+Simply repeating a previous motion a second time may not produce the desired effect.
 
@@
 
-Both demonstration and teleop suffer the limitation that a human operator must be immediately in the loop while the robot is operating.
-To make robots useful in the home, they will need to function {\em autonomously}, which leads to the notion of robot programming: giving high-level directives to robots so that they can function on their own without direct human guidance.
+Teleop suffers from the limitation that a human operator must be immediately in the loop while the robot is operating.
+To make robots useful in the home, they will need to function {\em autonomously}, which leads to the notion of end user programming: giving high-level directives to robots so that they can function on their own without direct human guidance.
 A full programming solution will likely involve a blend of different techniques including demonstration, teleop, and autonomous subroutines.
@@
 
-For personal robotics to truly succeed, robot programming will need to be simple enough that non-experts can do it.
+For personal robotics to truly succeed, regular people will need to use robot programming interfaces to tell robots what to do.
+These interfaces for robot programming will need to be simple enough for ordinary consumers to be able to pick up and use.
+In this article I detail some of the challenges that make robot control such an inherently difficult problem.
 
@@
 \begin{figure*}[tbh]
-\center\scalebox{0.20}{\includegraphics{rviz.png}}
-\caption{RViz interface for robot visualization}
+\center\scalebox{0.13}{
+	\includegraphics{rviz1.png}
+	\includegraphics{rviz2.png}
+}
+\caption{RViz interfaces for robot visualization}
 \label{rviz}
@@
 However, ROS provides a fairly low-level interface to robot control.
-Designed for roboticists, the RViz interface (Figure~\ref{rviz}) displays all the sensor feeds coming from the robot and lets one send low-level motion commands to the robot such as rolling forward or rotating the shoulder joint.
-In the figure, a robot is shown examining objects on a table in front of it.
-The green pattern represents the data from a depth sensor mounted on the robot's head, whereas the purple line represents the walls detected by a laser rangefinder mounted on the robot's base.
-The tricolored axes display the various coordinate frames in use, and the gray floor displays the map the robot uses for navigation.
-(TODO: display costmaps, TF axes)
+Designed for roboticists, the RViz interface (Figure~\ref{rviz}) displays all the sensor feeds coming from the robot and lets one send low-level motion commands to the robot such as rolling forward or rotating its shoulder joint.
+The figure shows two different perspectives on the same robot examining objects on a table in front of it.
+The left figure shows a third-person view of the robot and the table, rendered on top of a two-dimensional map of its environment (the gray pattern in the background).
+The green overlays on top of the tabletop objects indicate that they are recognized by the system as objects that can be picked up.
+The yellow grid pattern indicates horizontal surfaces that are detected, such as the floor.
+The right figure shows the view as soon from the robot's head-mounted camera.
+The red/green/blue arrows are control markers that can be used to rotate and move the robot's gripper in all directions.
 
@@
 
+Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
+Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
+
 %There is a tradeoff between exposing these joint limitations to the programmer, which allows precise yet complex control, versus insulating a programmer from those considerations, which may lead to unpredictable behavior that is difficult to debug.
 
-Yet positioning an arm through direct manipulation or teleop is relatively easy compared to doing it programmatically.
+% Yet positioning an arm through direct manipulation or teleop is relatively easy compared to doing it programmatically.
 In order to program an arm, one must first specify {\em where} it should go.
@@
 Other reference frames are defined relative to stationary parts of the robot, such as its base (if it is on wheels), the head-mounted camera, or the right elbow joint.
-
-Sending an arm to a specific position is useful for robot tasks such as, for example, looking at what it is holding by moving its hand to half a meter in front of its face.
+% 
+% Sending an arm to a specific position is useful for robot tasks such as, for example, looking at what it is holding by moving its hand to half a meter in front of its face.
 This half-meter is most reliably specified in the reference frame relative to the robot's head.
@@
 The opposite problem, calculating the joint positions required to get the hand to a particular position, is called {\em inverse kinematics} and is typically solved using either analytical equations or heuristic approximation methods.
-
-Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
-Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
 
@@
 
-This past generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff.
+The current generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff.
 iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''.
@@
 And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.
+
+% Acknowledgements
+% Andreas Paepke
+% Kaijen Hsiao
+% Matei Ciocarlie
 
@@
 \end{document}
+
+% vim: set lbr:

### 2013-11-29 paper.tex
First pass post-MIT-talk

@@
 Software could automate repetitive tasks in computer use and free people from mundane work.
-My chosen field, end user programming, had the potential to enable millions of ordinary computer users to customize and adapt systems to their own needs.
+So instead I developed expertise in human-computer interaction and intelligent user interfaces.
+My chosen field, end user programming, has the potential to enable millions of ordinary computer users to customize and adapt systems to their own needs.
 
@@
 
-Yet the holy grail of personal robotics remains elusive.
-When will we have our own Rosie the Robot to take over housekeeping chores?
-When will disabled adults have robot helpers to assist them in activities of daily living?
-All of these advances in robotics still fall short of the goal of creating personal robots that can do whatever we want them to do.
+Yet the holy grail of personal service robots remains elusive.
+I believe there are three main reasons.
+People have a wide variety of tasks for service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
+They also want robots to perform these tasks in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
+Most importantly, these service robots must be easy to use by non-technical people.
 
-What I have found is that the problem is not that robots {\em cannot} do what we want them to do.
-The problem is that it is too difficult to {\em tell} robots what we want them to do.
-Even basic robot use entails some form of programming, because autonomous algorithms are not yet reliable and robust enough to complete useful tasks.
-Except in a few narrow domains where automation has become robust and practical, the mere act of telling a robot what to do involves a complex set of skills.
-
-Robot control can be roughly categorized into three distinct phases: direct demonstration, tele-operation (or teleop for short), and end user programming.
-
-In demonstrational interfaces, a human operator moves a robot through a series of positions to demonstrate the motion it should perform.
-The robot then continues to perform this motion unattended.
-This style of programming is common for industrial robots, which can be programmed to move items around on conveyor belts or do precise assembly tasks.
-Demonstrational interfaces are typically limited to repeating only what was previously demonstrated, performing the exact same motion down to the millimeter.
-While this is useful in controlled environments such as factories, typical home environments are much more fluid and changing.
-Simply repeating a previous motion a second time may not produce the desired effect.
-
-Teleop consists of synchronously controlling a robot using software and hardware interfaces from a computer which may or may not be co-located with the robot.
-For example, a PS3 joystick controller could be used to drive a robot around a building, and activate certain pre-programmed motions such as turning its head or extending its arms.
-Remote teleop consists of controlling the robot over a distance, perhaps over the internet, using GUIs that allow the human operator to move various parts of the robot.
-
-Teleop suffers from the limitation that a human operator must be immediately in the loop while the robot is operating.
-To make robots useful in the home, they will need to function {\em autonomously}, which leads to the notion of end user programming: giving high-level directives to robots so that they can function on their own without direct human guidance.
-A full programming solution will likely involve a blend of different techniques including demonstration, teleop, and autonomous subroutines.
-For example, a task of setting the dinner table may include a high-level program of fetching silverware from the kitchen and placing it on the dining table, a demonstration of how to open this particular silverware drawer, and teleoperation when something goes awry (such as the silverware drawer being empty).
-
-For personal robotics to truly succeed, regular people will need to use robot programming interfaces to tell robots what to do.
-These interfaces for robot programming will need to be simple enough for ordinary consumers to be able to pick up and use.
-In this article I detail some of the challenges that make robot control such an inherently difficult problem.
+We have started to see robots enter the marketplace that address two out of those three hurdles.
+For example, iRobot's Roomba only does vacuuming, but it does it in a variety of environments and is dead simple to use even by non-technical folks.
+On the other hand, Rethink Robotics' Baxter robot can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
+As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
 
@@
 One important use case for personal robotics will be in-home care and assistance for disabled adults.
-The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aims to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
-Its first prototypes have enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
+The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aimed to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
+Its first prototypes enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
 Through physical therapy, Henry has regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
@@
 
-With this goal in mind, autonomous programmable robots are more than a consumer toy, but a technology that could changes people's lives for the better.
-Why, then, does robot programming for the masses remain such a difficult challenge?
+With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
+Why, then, are robots still so difficult to use?
 What makes robots so difficult to control?
+
+\section{Robot use is programming}
+
+Wikipedia defines computer programming as follows (italicized text added for clarification):
+
+\begin{quote}
+The purpose of programming is to find a sequence of instructions that will {\it [cause the computer to]} automate performing a specific task or solve a given problem.
+\end{quote}
+
+If you replace computer with robot, we come to a definition of robot programming: finding the right series of instructions that cause the robot to automate a task or exhibit a certain behavior.
+Every robot makes available a certain set of functionality -- its user-visible API.
+These are the primitive behaviors that have been programmed into the robot and can be activated by its user.
+End users must select which behavior to call when, in order to make the robot do what they want.
+Anyone who operates a robot, including consumers who purchase one for their home, will be programming the robot according to this definition.
+
+The difficulty of this programming task depends on several factors:
+
+\begin{itemize}
+\item {\bf Direct contact with the robot/task}: programming is easier if the user can see and touch the robot and its desired task.
+Control from a distance makes the programming process harder.
+For example, robots that are sent into natural disaster zones as first responders may have to be operated from a distance, which increases the programming task's difficulty.
+
+\item {\bf Task/environment variance}: if the task to be performed is always the same, the programming task is easier than if the goal task changes significantly or if the environment in which the robot must operate is changing.
+The program must be generalized to handle this variance, which makes the programming process more difficult.
+
+\item {\bf Goal specification}: some tasks are easy to describe, such as navigation goals specified by a single point in 3-dimensional space.
+Other goals, such as washing the dishes or tidying up, are more difficult to specify.
+Successful completion may be more dependent on how exactly the robot performs the task.
+Generally speaking, the more difficult it is to frame the goal specification, the more difficult the programming task will be.
+
+\item {\bf Amount of autonomy}: robots are gaining the ability to perform some functions autonomously, such as spatial navigation and object manipulation.
+The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior.
+Teleoperation interfaces are fairly easy to program because there is generally a one-to-one mapping between the user's inputs and the robot's behavior.
+Once a robot starts to act autonomously, understanding its behavior (and debugging it!) becomes more challenging.
+\end{itemize}
+
+For personal robotics to truly succeed, regular people will need to use robot programming interfaces to tell robots what to do.
+These interfaces for robot programming will need to be simple enough for ordinary consumers to be able to pick up and use.
+In this article I detail some of the challenges that make robot control such an inherently difficult problem.
 
@@
 }
-\caption{RViz interfaces for robot visualization}
+\caption{RViz interfaces for robot visualization and control}
 \label{rviz}
@@
 % Matei Ciocarlie
+% Jon Binney
+% Maya Cakmak
+% anonymous reviewers
 

### 2013-11-30 paper.tex
First pass at rewrite post-talk

@@
 \usepackage{graphicx}
+\usepackage{changebar}
 
@@
 \author {Tessa Lau \\
-Affiliation TBD \\
+Savioke Inc.\\
 Buenos Aires, Argentina \\
@@
 
+\begin{changebar}
 Yet the holy grail of personal service robots remains elusive.
@@
 As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
+\end{changebar}
 
@@
 With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
-Why, then, are robots still so difficult to use?
-What makes robots so difficult to control?
+\begin{changebar}
+Why, then, are robots still so difficult to use? What makes robots so difficult to control?
 
@@
 \end{itemize}
+\end{changebar}
 
@@
 
-This interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly.
+\begin{changebar}
+\section{Why robot programming is hard}
+\end{changebar}
+
+The Rviz interface, while well suited for roboticists who need to examine the low-level sensor data and messages being processed by the robot, is not very user-friendly.
 It also illustrates some of the difficulties inherent in robot programming.
 
-\section{A 3D world}
+\subsection{Concurrency}
+
+% ROS nodes
+% multiple parts of robots (head, arms, base)
+% callbacks, coroutines, threads
+
+\begin{changebar}
+Many activities of daily living require coordinated action amongst multiple limbs.
+Anyone who has lived with a broken arm, for example, understands the difficulty of trying to unscrew jar lids, pick up bulky objects, or chop vegetables with only a single arm.
+Controlling both arms simultaneously to perform coordinated actions is a matter of concurrent programming.
+
+The problem becomes even more complex with robots that have more moving parts.
+\end{changebar}
+For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
+Robots are also equipped with a slew of sensors including cameras, depth sensors, laser rangefinders, touch sensors, and feedback about its electrical and motor systems.
+These sensors and actuators must be operated together in coordination to produce a desired robotic behavior.
+
+\begin{changebar}
+Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
+Professional programmers learn to use advanced techniques such as multi-threading, event-driven callbacks, and coroutines in order to specify concurrent behavior in programs.
+Yet these concepts are notoriously difficult to use and debug, even for trained programmers; how can we expect non-technical users to be able to do the same?
+
+In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level.
+For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which should be scheduled in concert with foreground tasks (such as fetching a drink).
+\end{changebar}
+
+%Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
+For many tasks this is a reasonable solution.
+However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
+
+%End user programming systems have tried to simplify the problem of concurrency.
+Some systems visualize the different components on a timeline.
+This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
+A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
+More research is needed to discover what paradigm works best for controlling robot parts concurrently.
+
+% A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
+For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
+As another example, a two-armed robot could have both arms linked in a mirror configuration so that a command to move one arm would cause the other to mirror its motion, thus enabling the robot to pick up symmetric objects with two hands.
+
+\subsection{Uncertainty}
+
+% World changes out from under you
+% Hardware fails
+% Incomplete world models
+% Bayesian methods, probabilities
+
+Another challenge that robot programmers face is specifying how a robot should behave in an uncertain world.
+The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around.
+People walk around, objects get lost, furniture gets moved.
+Moreover, limitations in current sensing technology make it difficult to build a perfectly accurate model of the world.
+For example, differences in lighting make the same object look very different to an RGB camera.
+Therefore a robot cannot necessarily trust what it perceives, and it must be able to deal with the possibility that what it thinks is true is not actually true.
+
+In addition, robot hardware itself often produces unpredictable results.
+Unlike software, mechanical parts can slip, burn out, or break.
+They take time to actuate and do not come to a stop immediately.
+Giving a robot the same instruction multiple times may or may not result in the same outcome each time.
+
+Thus robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware.
+Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
+If the operator is not physically co-located with the robot, she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
+For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
+Should it simply try again?
+Move a little bit to get a better view and then try again?
+Ask a human?
+
+\begin{changebar}
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seem like an heavy burden for casual programmers who just want a robot to do something useful in the home.
+\end{changebar}
+
+%Most end user programming systems to date simply ignore the problem of uncertainty.
+For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.
+
+\begin{changebar}
+\subsection{Commonsense reasoning}
+
+Reasoning about objects in the world requires some amount of common sense which has yet to be programmed consistently into robots.
+For example, a robot who has been asked to pick up and deliver a coffee cup may well select a grasp that involves dipping its fingers into the liquid and picking up the cup by its side wall.
+Although the grasp would be valid, this behavior would be judged somewhat undesireable by human observers.
+
+As another example, robots that are capable of automated navigation may fail to observe human norms when navigating through human spaces.
+The custom of staying to the right when passing in a hallway, for example, can cause confusion if a robot charges down the middle of the hall and does not yield to oncoming passerby.
+
+Endowing computers with common sense is still an active research area.
+Techniques for making use of common sense reasoning have not yet made it into the vocabulary of professional programmers.
+Yet these concepts will be taken for granted by end users of robots, who will expect their robots to obey common social norms as they perform their tasks.
+\end{changebar}
+
+\subsection{A 3D world}
 
@@
 
-\section{7-DOF arms}
-
-A different aspect of high-dimensionality comes from the design of robot arms.
+%\section{7-DOF arms}
+%
+%A different aspect of high-dimensionality comes from the design of robot arms.
 A solid object in 3D space has six degrees of freedom (DOF): three dimensions to specify its position, and three more to specify its rotation.
@@
 That extra degree of freedom enables your arm to take on all those different shapes while still achieving the desired grasp.
-
-Interfaces for robot arm teleoperation typically involve setting the position or angle of each joint in the arm individually.
+%
+%Interfaces for robot arm teleoperation typically involve setting the position or angle of each joint in the arm individually.
 For example, a tablet-style interface might have fourteen buttons, one pair used to increase or decrease the angle of each joint in the arm.
@@
 Additionally, since the allowable joint positions are not completely independent, finding a correct solution is often done through trial and error.
-
-Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
+%
+%Toolkits for doing automatic motion planning are maturing, which will simplify the problem of robot arm control from how the arm should move in fine-grained detail to specifying only the high-level target position and orientation.
 Yet the challenge of specifying that 6-DOF target position and orientation still remains, and this will continue to be a challenge for easy-to-use robot interfaces that work with existing 2D displays and input devices.
+%
 
@@
 
-\section{Uncertainty}
-
-% World changes out from under you
-% Hardware fails
-% Incomplete world models
-% Bayesian methods, probabilities
-
-Another challenge that robot programmers face is specifying how a robot should behave in an uncertain world.
-The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around.
-People walk around, objects get lost, furniture gets moved.
-Moreover, limitations in current sensing technology make it difficult to build a perfectly accurate model of the world.
-For example, differences in lighting make the same object look very different to an RGB camera.
-Therefore a robot cannot necessarily trust what it perceives, and it must be able to deal with the possibility that what it thinks is true is not actually true.
-
-In addition, robot hardware itself often produces unpredictable results.
-Unlike software, mechanical parts can slip, burn out, or break.
-They take time to actuate and do not come to a stop immediately.
-Giving a robot the same instruction multiple times may or may not result in the same outcome each time.
-
-What does this mean for robot programmers?
-Robot programs must have mechanisms in place to deal with uncertain truth and unpredictable hardware.
-Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
-Unlike with demonstration and teleop, where the operator can see if the robot carried out the desired motion and take appropriate action, a robot programmer must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
-For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
-Should it simply try again?
-Move a little bit to get a better view and then try again?
-Ask a human?
-
-In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and learn to craft fault-tolerant code with frequent error checking, which seems like an heavy burden for casual programmers.
-Most end user programming systems to date simply ignore the problem of uncertainty.
-For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.
-
-\section{Concurrency}
-
-% ROS nodes
-% multiple parts of robots (head, arms, base)
-% callbacks, coroutines, threads
-
-Robots are complex beings full of many moving parts.
-For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
-Robots are also equipped with a slew of sensors including cameras, depth sensors, laser rangefinders, touch sensors, and feedback about its electrical and motor systems.
-Sensors and actuators must be operated together in coordination to produce a desired robotic behavior.
-
-Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
-Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
-For many tasks this is a reasonable solution.
-However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
-
-End user programming systems have tried to simplify the problem of concurrency.
-Some systems visualize the different components on a timeline.
-This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
-A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
-More research is needed to discover what paradigm works best for controlling robot parts concurrently.
-
-% A second option for concurrency is to reduce the end user's need to write concurrent programs by defining new primitives that cover the common cases where concurrency is needed.
-For example, a robot navigation algorithm may include the ability to turn the robot's head to look in the direction of motion, eliminating the need for the user to specify this complex behavior.
-As another example, a two-armed robot could have both arms linked in a mirror configuration so that a command to move one arm would cause the other to mirror its motion, thus enabling the robot to pick up symmetric objects with two hands.
-
-Even if individual robot parts can be programmed to act in parallel, there will still be a need for reasoning about concurrent robot behavior at a higher level.
-One form this will take is for background tasks, such as specifying that a robot should try to tidy up the room in between (or during) its other assigned chores.
-Supporting high level concurrency will introduce new problems, such as task prioritization (is the current task of bringing you a beer more important than cleaning up the wine spill it just detected?).
-
 \section{The way forward}
 
-Robot programming environments are beginning to address all of these challenges and produce useful systems that can be controlled by untrained users.
-The key tradeoff is simplicity versus expressiveness: in order to make systems that are more tractable for novices, they need to simplify key aspects of the robot's behavior, thereby sacrificing its ability to perform a full range of behavior.
+\begin{changebar}
+The field of end user programming~\cite{lieberman-yourwish,dontcheva-nocode} has developed a variety of solutions to enable users of software systems to create or modify software artifacts without having to learn the intricacies of programming.
 
-The current generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff.
+The key insight for these EUP systems is to reduce the complexity of a programming task through simplification.
+The three primary methods include (a) visual programming: simplifying or eliminating tricky syntax through graphical representations; (b) domain-specific languages: limiting concepts to only those necessary in a particular domain of use, such as spreadsheet formulas; and (c) programming by demonstration or example: letting users focus on concrete examples rather than difficult abstractions.
+
+However, this simplicity typically comes at the price of expressiveness.
+By reducing the complexity exposed in an interface, we also reduce the number or variety of functions that can be performed in that interface.
+In order to make robots that are more tractable for non-technical users, we will have to simplify key aspects of robot behavior, thus sacrificing robots' ability to perform a full range of behavior.
+\end{changebar}
+
+%The current generation of commercial home robots illustrates one extreme end of the simplicity versus expressiveness tradeoff.
 iRobot's successful Roomba line of vacuum-cleaning robots\footnote{http://www.irobot.com/us/learn/home/roomba.aspx} sports a single-button interface that simply says ``Clean''.
@@
 
-Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles)\footnote{MoveIt!, http://moveit.ros.org}, and recognizing and picking up discrete objects on a flat surface.
+%Roboticists are hard at work developing increasingly autonomous robot capabilities for well-specified tasks such as 2D navigation (getting from point A to point B within a building), arm motion planning (moving the gripper to a specified position and orientation while avoiding obstacles)\footnote{MoveIt!, http://moveit.ros.org}, and recognizing and picking up discrete objects on a flat surface.
 Autonomy is key for enabling robots to automatically perform longer and more complex tasks on their own.
 
-Our approach to designing end-user-programmable interfaces is to craft these autonomous components into a good set of {\em primitives} in an end user programming language, such that a non-expert would not have to ``look inside the box'' to understand how each one works.
+\begin{changebar}
+\subsection{Higher-level primitives}
+
+One way to trade off expressiveness for simplicity is to abstract away details of robot behavior into higher-level primitives.
+This abstraction can be difficult for roboticists to accept, who have become accustomed to having fine-grained control over robot behavior and error handling.
+Yet abstracting away these details is going to be critical for end users to be able to develop a working understanding of the robot's functionality.
+
+%Our approach to designing end-user-programmable interfaces is to craft these autonomous components into a good set of {\em primitives} in an end user programming language, such that a non-expert would not have to ``look inside the box'' to understand how each one works.
+
 Roboticists can then implement these primitives using advanced techniques to solve the difficult problems encountered during robot programming.
+\end{changebar}
 For example, a good autonomous navigation component could incorporate a set of recovery behaviors to try on navigation failure, such as turning around, waiting for people/obstacles to clear out of its path, and backing up and retrying.
@@
 
-We have made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enables casual users to control our PR-2 robots through a web interface (Figure~\ref{contops}).
-This web interface, accessible from standard laptops or tablets, provides the ability to tell a PR-2 robot to unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
+We made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enabled casual users to control PR-2 robots through a web interface (Figure~\ref{contops}).
+\begin{changebar}
+The interface was inspired by the head of Willow Garage, who noted that even though he had a lab full of robots, he himself wasn't able to show off any of the functionality of his robots to visitors without a roboticist present to operate them.
+\end{changebar}
 
-This interface lies somewhere between teleoperation and programming.
-It does not require an operator to exert direct control over the robot, but the commands that are given are executed immediately by the robot rather than scheduled to occur later in time.
-We are also exploring interfaces for sequencing these primitives into scheduled programs, in order to validate that the set of primitives we are developing is complete enough to perform useful tasks.
+In response, we developed a web interface, accessible from standard laptops or tablets, that provided the ability to tell a PR-2 robot to perform a variety of tasks.
+It could unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
+\begin{changebar}
+The interface enabled non-technical users to activate several core autonomous functions of the PR-2 through an easy-to-use interface.
 
-Another key component of our approach, which we call {\em shared autonomy}, is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future.
-In our system designs, the teleoperator does not necessarily have to be the person who wrote the program.
-Instead, we are modeling our system after telephone call centers where a team of trained professionals handle incidents as they arise and get the robots back on track to fulfill their programmed goals.
-As a stepping-stone towards full autonomy, this approach enables us to start testing end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.
+The drawback of this interface was that the user lost fine-grained control over exactly how the robot performed these tasks.
+For example, the interface only supported using the robot's right arm to pick up an object; the left arm was unused.
+It did not allow the user to specify exactly how to pick up an object (e.g.
+from above or from the side), even though these parameters can be customized at a lower level of control.
+The challenge in designing a simpler interface is to carefully craft exactly which functionality to expose, and which functionality can be safely hidden from casual users.
 
-Over time, as we refine existing autonomous primitives and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior.
+\subsection{Shared autonomy}
+
+Another way to simplify the programming problem for end users is to take advantage of {\em shared autonomy}.
+The idea behind shared autonomy is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future.
+A teleoperator does not necessarily have to be the person who wrote the program.
+Instead, they could function similarly to staff in telephone call centers, where a team of trained professionals handle incidents as they arise and get the robots back on track to fulfill their programmed goals.
+
+With a human expert on call to get a robot out of situations it was not programmed to handle, we can simplify the programming problem for end users so that they do not have to specify error recovery behaviors for every single corner case.
+\end{changebar}
+
+As a stepping-stone towards full autonomy, this approach also enables the testing of end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.
+
+\begin{changebar}
+\subsection{New 3D input devices}
+
+Finally, video gaming technology is driving a new crop of input and output devices that make it easier to interact with and perceive the 3D world around us.
+These devices will make it easier for robots to perceive the world in three dimensions, and for people to instruct those robots using more natural gestures.
+Input devices such as the Razer Hydra, Leap Motion Controller, and Microsoft Kinect can sense motion in 3D, enabling computers to sense motion and objects in the world.
+3D output devices such as the Oculus Rift project stereo images in 3D so that people can perceive what robots are seeing in three dimensions.
+
+These new devices could enable a new generation of robot control interfaces that make interacting with robots more natural and intuitive.
+
+\section{Conclusion}
+
+Personal service robots have an enormous potential to do good for humanity by enabling people to live more independently.
+However, controlling these robots boils down to a small matter of programming.
+In this article I have explained why programming robots is inherently difficult, even for skilled programmers.
+Drawing lessons from the field of end user programming, we have proposed several paths forward that may enable the creation of more accessible interfaces for personal robots in the future.
+\end{changebar}
+
+%Over time, as we refine existing autonomous primitives and build new domain-specific primitives, my hope is that we will expand the repertoire of robust end-user-programmable robot behavior.
 And perhaps that goal of having our own Rosie the Robot in our homes will soon become a reality.
@@
 % Jon Binney
+% Brandon Alexander
 % Maya Cakmak

### 2013-12-01 paper.tex
Final read through first half (up to EUP section)

@@
 I believe there are three main reasons.
-People have a wide variety of tasks for service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
-They also want robots to perform these tasks in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
-Most importantly, these service robots must be easy to use by non-technical people.
+There is a wide variety of tasks that people want service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
+These tasks must be performed in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
+Most importantly, the robots themselves must be easy to operate by non-technical people.
 
-We have started to see robots enter the marketplace that address two out of those three hurdles.
-For example, iRobot's Roomba only does vacuuming, but it does it in a variety of environments and is dead simple to use even by non-technical folks.
-On the other hand, Rethink Robotics' Baxter robot can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
+We have started to see robots enter the marketplace that address two out of those three challenges.
+For example, iRobot's Roomba\footnote{XXX iRobot website} only vacuums floors, but it does so in a variety of environments and is dead simple to use even by non-technical folks.
+On the other hand, Rethink Robotics' Baxter robot\footnote{XXX Baxter website} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
 As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
@@
 
-One important use case for personal robotics will be in-home care and assistance for disabled adults.
+One important use case for personal service robots will be in-home care and assistance for disabled adults.
 The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aimed to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
@@
 Through physical therapy, Henry has regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
-Using a teleop interface custom designed for him, Henry has been able to instruct Willow Garage's PR-2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+Using an interface custom designed for him, Henry has been able to instruct Willow Garage's PR-2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
 
@@
 While Henry has only limited interaction capabilities, he is also strongly motivated because he has no alternatives.
-Able-bodied consumers, on the other hand, would quickly lose patience.
-If even basic robot control via teleoperation is so time-consuming, what hope do we have that people will ever get around to the general task of robot programming?
+\begin{changebar}
+Able-bodied consumers, on the other hand, would quickly lose patience and simply fetch the towel themselves.
+If the state of the art in robot operation is so time-consuming, what hope do we have that ordinary mortals will be able to operate robots to perform these tasks in their own homes?
+\end{changebar}
 
@@
 \begin{changebar}
-Why, then, are robots still so difficult to use? What makes robots so difficult to control?
+Why, then, are robots still so difficult to use? What makes robots so difficult to operate?
 
@@
 
-If you replace computer with robot, we come to a definition of robot programming: finding the right series of instructions that cause the robot to automate a task or exhibit a certain behavior.
+If you replace computer with robot, we come to a definition of robot programming: finding the right series of instructions that cause a robot to automate a task or exhibit a certain behavior.
 Every robot makes available a certain set of functionality -- its user-visible API.
@@
 Control from a distance makes the programming process harder.
-For example, robots that are sent into natural disaster zones as first responders may have to be operated from a distance, which increases the programming task's difficulty.
+For example, robots that are sent into natural disaster zones as first responders may have to be operated from a distance, which increases the programming task's difficulty because the operator's senses are limited to what can be perceived through the robot's own cameras or sensors.
 
@@
 Many activities of daily living require coordinated action amongst multiple limbs.
-Anyone who has lived with a broken arm, for example, understands the difficulty of trying to unscrew jar lids, pick up bulky objects, or chop vegetables with only a single arm.
+Anyone who has broken an arm, for example, understands the difficulty of trying to unscrew jar lids, pick up bulky objects, or chop vegetables with only a single arm.
 Controlling both arms simultaneously to perform coordinated actions is a matter of concurrent programming.
 
-The problem becomes even more complex with robots that have more moving parts.
+The problem becomes even more complex with robots that have many moving parts.
 \end{changebar}
@@
 Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
-If the operator is not physically co-located with the robot, she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
+\begin{changebar}
+If an operator is not physically co-located or co-temporaneous with the robot,
+\end{changebar}
+she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
 For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
@@
 \begin{changebar}
-In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seem like an heavy burden for casual programmers who just want a robot to do something useful in the home.
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seems like too much to ask of casual programmers who just want a robot to do something useful in the home.
 \end{changebar}
@@
 
-\section{The way forward}
+\begin{changebar}
+\section{End user programming for robots}
 
-\begin{changebar}
 The field of end user programming~\cite{lieberman-yourwish,dontcheva-nocode} has developed a variety of solutions to enable users of software systems to create or modify software artifacts without having to learn the intricacies of programming.

### 2013-12-02 paper.tex
Revision complete, sending for Steve's approval

@@
 \maketitle
-
-% Feedback from Matei:
-
-% Make assumptions more explicit in intro
-% Robot use == robot programming
-% Differentiate end user programming from roboticist programming
-% Teach/instruct/tell == programming
-
-% Non-quantitative Henry statistics
-
-% Cartesian robot control is a solved problem
 
@@
 This abstraction can be difficult for roboticists to accept, who have become accustomed to having fine-grained control over robot behavior and error handling.
-Yet abstracting away these details is going to be critical for end users to be able to develop a working understanding of the robot's functionality.
+Yet abstracting away these details is going to be critical for end users to be able to develop a working understanding of a robot's functionality.
 
@@
 from above or from the side), even though these parameters can be customized at a lower level of control.
+
 The challenge in designing a simpler interface is to carefully craft exactly which functionality to expose, and which functionality can be safely hidden from casual users.
+For the intended users of the Continuous Ops interface, the provided functionality was sufficient and satisfied the requirements of being easy to use and always available.
+In fact, we heard of several cases where people used the interface remotely (e.g.
+while away at a conference) to show off the functionality of the robots in the lab.
 
@@
 \begin{changebar}
+This approach worked well during a multi-day deployment of a delivery system at Willow Garage.
+Users could request delivery of chocolates or snacks to their office, and the system would autonomously retrieve those items and deliver them to their office door.
+In the rare cases that the autonomous behaviors failed, several roboticists were notified by email and played the role of teleoperators.
+They initiated recovery behaviors to get the robots back on track.
+From the end users' perspective, the robot performed its tasks fully autonomously, sparing them from having to worry about specifying recovery behaviors.
+\end{changebar}
+
+\begin{changebar}
 \subsection{New 3D input devices}
@@
 These devices will make it easier for robots to perceive the world in three dimensions, and for people to instruct those robots using more natural gestures.
-Input devices such as the Razer Hydra, Leap Motion Controller, and Microsoft Kinect can sense motion in 3D, enabling computers to sense motion and objects in the world.
+Input devices such as the Razer Hydra, Leap Motion Controller, and Microsoft Kinect can sense motion in 3D, enabling robots to better sense motion and objects in the world.
 3D output devices such as the Oculus Rift project stereo images in 3D so that people can perceive what robots are seeing in three dimensions.
@@
 These new devices could enable a new generation of robot control interfaces that make interacting with robots more natural and intuitive.
+For example, David Gossow at Willow Garage developed a research prototype called the PR-2 Surrogate which provided a fully-immersive teleoperation interface for the PR-2 robot.
+It used a Razer Hydra control to track your hand positions, which were then mapped directly to the robot's hand positions.
+It also displayed a stereo camera feed from the robot's head-mounted cameras into an Oculus Rift device, giving you a stereo view of what the robot was seeing.
+The Rift's head tracking capability was mapped to moving the robot's head, enabling you to look around the robot's environment just by turning your head.
+This demo showed that these new input/output technologies hold much promise for making it easier to interact with, control, and program complex robots.
 
@@
 
-Personal service robots have an enormous potential to do good for humanity by enabling people to live more independently.
+Personal service robots have an enormous potential to do good for humanity by enabling people to live more independent lives and free them from mundane housekeeping chores.
 However, controlling these robots boils down to a small matter of programming.
 In this article I have explained why programming robots is inherently difficult, even for skilled programmers.
-Drawing lessons from the field of end user programming, we have proposed several paths forward that may enable the creation of more accessible interfaces for personal robots in the future.
+Drawing lessons from the field of end user programming, I have proposed several paths forward that may enable the creation of more accessible interfaces for personal robots in the future.
 \end{changebar}

### 2014-08-17 paper.tex
Removed changebars from last submission

@@
+% vim: set lbr:
 \documentclass[10pt,twocolumn]{article}
@@
 
-\begin{changebar}
 Yet the holy grail of personal service robots remains elusive.
@@
 We have started to see robots enter the marketplace that address two out of those three challenges.
-For example, iRobot's Roomba\footnote{XXX iRobot website} only vacuums floors, but it does so in a variety of environments and is dead simple to use even by non-technical folks.
-On the other hand, Rethink Robotics' Baxter robot\footnote{XXX Baxter website} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
+For example, iRobot's Roomba\footnote{http://www.irobot.com} only vacuums floors, but it does so in a variety of environments and is dead simple to use even by non-technical folks.
+On the other hand, Rethink Robotics' Baxter robot\footnote{http://www.rethinkrobotics.com/products/baxter/} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
 As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
-\end{changebar}
 
@@
 While Henry has only limited interaction capabilities, he is also strongly motivated because he has no alternatives.
-\begin{changebar}
 Able-bodied consumers, on the other hand, would quickly lose patience and simply fetch the towel themselves.
 If the state of the art in robot operation is so time-consuming, what hope do we have that ordinary mortals will be able to operate robots to perform these tasks in their own homes?
-\end{changebar}
 
@@
 With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
-\begin{changebar}
 Why, then, are robots still so difficult to use? What makes robots so difficult to operate?
@@
 \end{itemize}
-\end{changebar}
 
@@
 
-\begin{changebar}
 \section{Why robot programming is hard}
-\end{changebar}
 
@@
 
-\begin{changebar}
 Many activities of daily living require coordinated action amongst multiple limbs.
@@
 The problem becomes even more complex with robots that have many moving parts.
-\end{changebar}
 For example, Willow Garage's PR-2 has two independently controlled 7-DOF arms, a head that can pan and tilt, a torso that can be raised and lowered, and a base that can move and turn in all directions.
@@
 
-\begin{changebar}
 Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
@@
 For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which should be scheduled in concert with foreground tasks (such as fetching a drink).
-\end{changebar}
 
@@
 Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
-\begin{changebar}
 If an operator is not physically co-located or co-temporaneous with the robot,
-\end{changebar}
 she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
@@
 
-\begin{changebar}
 In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seems like too much to ask of casual programmers who just want a robot to do something useful in the home.
-\end{changebar}
 
@@
 
-\begin{changebar}
 \subsection{Commonsense reasoning}
@@
 Yet these concepts will be taken for granted by end users of robots, who will expect their robots to obey common social norms as they perform their tasks.
-\end{changebar}
 
@@
 
-\begin{changebar}
 \section{End user programming for robots}
@@
 In order to make robots that are more tractable for non-technical users, we will have to simplify key aspects of robot behavior, thus sacrificing robots' ability to perform a full range of behavior.
-\end{changebar}
 
@@
 
-\begin{changebar}
 \subsection{Higher-level primitives}
@@
 Roboticists can then implement these primitives using advanced techniques to solve the difficult problems encountered during robot programming.
-\end{changebar}
 For example, a good autonomous navigation component could incorporate a set of recovery behaviors to try on navigation failure, such as turning around, waiting for people/obstacles to clear out of its path, and backing up and retrying.
@@
 We made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enabled casual users to control PR-2 robots through a web interface (Figure~\ref{contops}).
-\begin{changebar}
 The interface was inspired by the head of Willow Garage, who noted that even though he had a lab full of robots, he himself wasn't able to show off any of the functionality of his robots to visitors without a roboticist present to operate them.
-\end{changebar}
 
@@
 It could unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
-\begin{changebar}
 The interface enabled non-technical users to activate several core autonomous functions of the PR-2 through an easy-to-use interface.
@@
 With a human expert on call to get a robot out of situations it was not programmed to handle, we can simplify the programming problem for end users so that they do not have to specify error recovery behaviors for every single corner case.
-\end{changebar}
 
@@
 
-\begin{changebar}
 This approach worked well during a multi-day deployment of a delivery system at Willow Garage.
@@
 From the end users' perspective, the robot performed its tasks fully autonomously, sparing them from having to worry about specifying recovery behaviors.
-\end{changebar}
 
-\begin{changebar}
 \subsection{New 3D input devices}
@@
 Drawing lessons from the field of end user programming, I have proposed several paths forward that may enable the creation of more accessible interfaces for personal robots in the future.
-\end{changebar}
 
@@
 \end{document}
-
-% vim: set lbr:

### 2014-08-18 paper.tex
Started adding some refs

@@
-% vim: set lbr:
 \documentclass[10pt,twocolumn]{article}
@@
 I believe there are three main reasons.
-There is a wide variety of tasks that people want service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
-These tasks must be performed in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
-Most importantly, the robots themselves must be easy to operate by non-technical people.
+First, there is a wide variety of tasks that people want service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
+Second, these tasks must be performed in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
+Finally and most importantly, the robots themselves must be easy to operate by non-technical people.
 
@@
 
-Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does.
+Concurrent programming is notoriously difficult even for professional programmers due to the possibility for race conditions and the complexity of reasoning about what the program does~\cite{shavit-cacm11}.
 Professional programmers learn to use advanced techniques such as multi-threading, event-driven callbacks, and coroutines in order to specify concurrent behavior in programs.
@@
 
-In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level.
+In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level~\cite{cakmak-hri13}.
 For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which should be scheduled in concert with foreground tasks (such as fetching a drink).
@@
 Another challenge that robot programmers face is specifying how a robot should behave in an uncertain world.
-The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around.
+The world is constantly changing: any model that a robot could build of where things are located will be obsolete by the time it turns around~\cite{fox-jair99}.
 People walk around, objects get lost, furniture gets moved.
-Moreover, limitations in current sensing technology make it difficult to build a perfectly accurate model of the world.
+Moreover, limitations in current sensing technology make it difficult to build a perfectly accurate model of the world~\cite{kemp-ieee07}.
 For example, differences in lighting make the same object look very different to an RGB camera.
@@
 
-In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models and craft fault-tolerant code with frequent error checking, which seems like too much to ask of casual programmers who just want a robot to do something useful in the home.
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking, which seems like too much to ask of casual programmers who just want a robot to do something useful in the home.
 
@@
 
-Endowing computers with common sense is still an active research area.
+Endowing computers with common sense is still an active research area~\cite{mueller-cacm09}.
 Techniques for making use of common sense reasoning have not yet made it into the vocabulary of professional programmers.

### 2014-08-18 paper.tex
More references

@@
 Savioke Inc.\\
-Buenos Aires, Argentina \\
-{\tt tlau@tlau.org}
+Santa Clara, CA\\
+{\tt tlau@savioke.com}
 }
@@
 
-All these innovations have become possible through multiple advances: in sensor technology, which robots use to perceive the world around them; in planning and navigation, which enable robots to move around in the world unaided; in human-safe motion controllers, which let robots move their limbs without endangering the people around them; in the ROS open source robot operating system, which enables roboticists to build on each others' work without starting from scratch; and many more advances in the broader field of robotics.
+All these innovations have become possible through multiple advances: in sensor technology, which robots use to perceive the world around them~\cite{henry-iser10}; in planning and navigation, which enable robots to move around in the world unaided~\cite{jackel-cacm07}; in human-safe motion controllers, which let robots move their limbs without endangering the people around them~\cite{kemp-ieee07}; in the ROS open source robot operating system, which enables roboticists to build on each others' work without starting from scratch~\cite{quigley-icra09}; and many more advances in the broader field of robotics.
 

### 2014-08-25 paper.tex
Added a few more refs

@@
 We have started to see robots enter the marketplace that address two out of those three challenges.
-For example, iRobot's Roomba\footnote{http://www.irobot.com} only vacuums floors, but it does so in a variety of environments and is dead simple to use even by non-technical folks.
-On the other hand, Rethink Robotics' Baxter robot\footnote{http://www.rethinkrobotics.com/products/baxter/} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility make it less suitable for the home.
+For example, iRobot's Roomba\footnote{http://www.irobot.com} only vacuums floors, but it does so in a variety of environments and is simple to use even by non-technical folks.
+On the other hand, Rethink Robotics' Baxter robot\footnote{http://www.rethinkrobotics.com/products/baxter/} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility makes it less suitable for the home.
 As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
+
+Yet I believe the time is ripe for a breakthrough in personal service robotics, starting with robots that perform simpler tasks in relatively structured environments such as hospitals or hotels, and driven by better design and usable interfaces that enable them to be used by non-expert users.
 
@@
 One important use case for personal service robots will be in-home care and assistance for disabled adults.
-The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aimed to develop technology that enables persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
+The Robots for Humanity project~\cite{rfh}, a collaboration between Willow Garage and Georgia Tech, aimed to develop technology that enabled persons with motor impairments to perform personal care tasks that they otherwise might not have been able to perform without assistance, such as picking up a dropped object, fetching a towel, or even scratching an itch.
 Its first prototypes enabled Henry Evans, a mute quadriplegic, to act in the world for the first time since a brainstem stroke left him paralyzed.
-Through physical therapy, Henry has regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
-Using an interface custom designed for him, Henry has been able to instruct Willow Garage's PR-2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
+Through physical therapy, Henry had regained the ability to make limited head motions (enough to control a cursor with his gaze) and to click a button with one finger.
+Using an interface custom designed for him, Henry was able to instruct Willow Garage's PR-2 robots to open cabinets in his home, fetch food from the refrigerator, and even dole out candy to trick-or-treating kids on his behalf.
 
-Yet our experience with Robots for Humanity underscores how difficult it is for ordinary mortals to instruct robots what to do.
-Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door could be very difficult for Henry to accomplish.
-While Henry has only limited interaction capabilities, he is also strongly motivated because he has no alternatives.
+Yet our experience with Robots for Humanity underscored how difficult it was for ordinary mortals to instruct robots what to do.
+Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door was very difficult for Henry to accomplish.
+While Henry had only limited interaction capabilities, he was also strongly motivated because he has no alternatives.
 Able-bodied consumers, on the other hand, would quickly lose patience and simply fetch the towel themselves.
@@
 With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
-Why, then, are robots still so difficult to use? What makes robots so difficult to operate?
+Why, then, are robots still so difficult to use?
+What makes robots so difficult to operate?
 
-\section{Robot use is programming}
+\section{Robot programming}
 
@@
 End users must select which behavior to call when, in order to make the robot do what they want.
-Anyone who operates a robot, including consumers who purchase one for their home, will be programming the robot according to this definition.
+%Anyone who operates a robot, including consumers who purchase one for their home, will be programming the robot according to this definition.
 
@@
 \item {\bf Direct contact with the robot/task}: programming is easier if the user can see and touch the robot and its desired task.
-Control from a distance makes the programming process harder.
+Control from a distance makes the programming process harder~\cite{chen-ieeetos07}.
 For example, robots that are sent into natural disaster zones as first responders may have to be operated from a distance, which increases the programming task's difficulty because the operator's senses are limited to what can be perceived through the robot's own cameras or sensors.
@@
 The program must be generalized to handle this variance, which makes the programming process more difficult.
+% REF: something from the EUP literature? Andy Ko or Margaret Burnett?
 
@@
 Generally speaking, the more difficult it is to frame the goal specification, the more difficult the programming task will be.
+% REF: something from the EUP literature?
 
@@
 The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior.
-Teleoperation interfaces are fairly easy to program because there is generally a one-to-one mapping between the user's inputs and the robot's behavior.
+Teleoperation interfaces are fairly easy to use because there is generally a one-to-one mapping between the user's inputs and the robot's behavior.
 Once a robot starts to act autonomously, understanding its behavior (and debugging it!) becomes more challenging.

### 2014-08-26 paper.tex
More references

@@
 
-The difficulty of this programming task depends on several factors:
+Many of the lessons learned from the fields of end user programming~\cite{dontcheva-nocode} and end user software engineering~\cite{ko-acm11} also apply to robot programming, though robots' physical embodiment leads to additional challenges.
+I believe that the difficulty of the robot programming task boils down to several factors:
 
@@
 Successful completion may be more dependent on how exactly the robot performs the task.
-Generally speaking, the more difficult it is to frame the goal specification, the more difficult the programming task will be.
+Generally speaking, the more difficult it is to frame the goal specification, the more difficult the programming task will be~\cite{ko-vlhcc04}.
 % REF: something from the EUP literature?
@@
 \item {\bf Amount of autonomy}: robots are gaining the ability to perform some functions autonomously, such as spatial navigation and object manipulation.
-The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior.
+The less autonomy exhibited by the robot, the easier it is to program the robot, because increased autonomy requires the programmer to develop more complex mental models of the robot's perceptions and behavior~\cite{kulesza-chi12}.
 Teleoperation interfaces are fairly easy to use because there is generally a one-to-one mapping between the user's inputs and the robot's behavior.

### 2014-08-28 paper.tex
More work

@@
 Many of the lessons learned from the fields of end user programming~\cite{dontcheva-nocode} and end user software engineering~\cite{ko-acm11} also apply to robot programming, though robots' physical embodiment leads to additional challenges.
-I believe that the difficulty of the robot programming task boils down to several factors:
+Based on my experience in both end user programming and robotics, I believe that the difficulty of the robot programming task boils down to several factors:
 
@@
 The Robot Operating System\cite{ros} is one of the software advances that supports the rapid development of new robotics technology.
-First developed in 2007 at the Stanford Artificial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has recently transitioned into the stewardship of the non-profit Open Source Robotics Foundation.
+First developed in 2007 at the Stanford Artificial Intelligence Laboratory, ROS was maintained by Willow Garage from 2008-2013 and has since transitioned into the stewardship of the non-profit Open Source Robotics Foundation\footnote{http://www.osrfoundation.org/}.
 
@@
 The right figure shows the view as soon from the robot's head-mounted camera.
-The red/green/blue arrows are control markers that can be used to rotate and move the robot's gripper in all directions.
+The red/green/blue arrows are control markers that can be used to rotate and move the robot's gripper in all directions~\cite{gossow-ieee11}.
 
@@
 Professional programmers learn to use advanced techniques such as multi-threading, event-driven callbacks, and coroutines in order to specify concurrent behavior in programs.
-Yet these concepts are notoriously difficult to use and debug, even for trained programmers; how can we expect non-technical users to be able to do the same?
+Yet these concepts are notoriously difficult to use and debug, even for trained programmers; how can we expect non-technical users to be able to do the same, in order to instruct multi-limbed robots?
 
 In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level~\cite{cakmak-hri13}.
-For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which should be scheduled in concert with foreground tasks (such as fetching a drink).
+For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which might be scheduled in concert with foreground tasks (such as fetching a drink).
 
@@
 
-%End user programming systems have tried to simplify the problem of concurrency.
+End user programming systems have tried various techniques to deal with the problem of concurrency.
 Some systems visualize the different components on a timeline.
-This approach is popular with media creation tools such as Apple's GarageBand or Autodesk's 3D Studio Max.
+This approach is popular with media creation tools such as Apple's GarageBand\footnote{https://www.apple.com/mac/garageband/} or Autodesk's 3D Studio Max\footnote{http://www.autodesk.com/products/3ds-max/overview}.
 A second paradigm, taken by visual programming languages such as Scratch~\cite{scratch}, makes the notion of time a global constant and runs all blocks of user code simultaneously in the virtual environment.
@@
 
-In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking, which seems like too much to ask of casual programmers who just want a robot to do something useful in the home.
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking, which seems like too much to ask of casual programmers who just to teach their robot to do something useful in the home.
 

### 2014-08-28 paper.tex
Work in progress

@@
 \maketitle
+
+% Hai Nguyen's ROCON
+% Maya's Robot PBD
+% Strips style robot planning
 
@@
 Generally speaking, the more difficult it is to frame the goal specification, the more difficult the programming task will be~\cite{ko-vlhcc04}.
-% REF: something from the EUP literature?
 
@@
 These interfaces for robot programming will need to be simple enough for ordinary consumers to be able to pick up and use.
-In this article I detail some of the challenges that make robot control such an inherently difficult problem.
-
-The explanation begins with a look at the state of the art in robot software today.
-
-% Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
+Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
 Why is end user robot programming so challenging?
 Several compounding factors make it more difficult than normal computer use.
+
+The story begins with a look at the state of the art in robot software today.
 
@@
 Robot control loops usually sense the physical hardware during operation to determine whether a motion was carried out successfully.
-If an operator is not physically co-located or co-temporaneous with the robot,
-she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
+If an operator is not physically co-located or co-temporaneous with the robot, she cannot see whether the robot carried out the desired motion, so she must think through possible outcomes while writing a program and craft appropriate responses for each eventuality.
 For example, while instructing a robot to pick up an object, a programmer must specify what to do if the object is not {\em where} it thinks it is, is not {\em what} it thinks it is, or if the robot fails to perform the desired grasp.
@@
 
-In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking, which seems like too much to ask of casual programmers who just to teach their robot to do something useful in the home.
+In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking.
+Nguyen's ROS Commander (ROSCo) system is an end-user programming system aimed to enable home users to define new behaviors for PR2 robots by visually crafting finite state machines~\cite{nguyen-icra13}.
+ROSCo XXX TODO
 
-%Most end user programming systems to date simply ignore the problem of uncertainty.
-For robot programming to be successful, we will need to develop easily-understandable paradigms for handling uncertainty.
+which seems like too much to ask of casual programmers who just want to teach their robot to do something useful in the home.
 

### 2014-08-28 paper.tex
More work

@@
 
-% Hai Nguyen's ROCON
 % Maya's Robot PBD
@@
 
-In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level~\cite{cakmak-hri13}.
-For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which might be scheduled in concert with foreground tasks (such as fetching a drink).
+In addition to the low level concurrency needed to control multiple robot parts, there is also a need to specify concurrent robot behavior at a higher level~\cite{cakmak-hri13}, or across multiple robots.
+For example, a user may want to give the robot background tasks (such as cleaning or plugging oneself in when the battery is low) which might be scheduled in concert with foreground tasks (such as fetching a drink), or might want to command two robots to work together to perform a task~\cite{liu-chi11}.
 
@@
 
-End user programming systems have tried various techniques to deal with the problem of concurrency.
+Other end user programming systems have tried various techniques to deal with the problem of concurrency.
 Some systems visualize the different components on a timeline.
@@
 In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking.
-Nguyen's ROS Commander (ROSCo) system is an end-user programming system aimed to enable home users to define new behaviors for PR2 robots by visually crafting finite state machines~\cite{nguyen-icra13}.
-ROSCo XXX TODO
-
-which seems like too much to ask of casual programmers who just want to teach their robot to do something useful in the home.
+Systems such as Nguyen's ROS Commander~\cite{nguyen-icra13} enable non-roboticists to define and customize robot behaviors, handling error conditions.
+However, ROS Commander's programming interface, which the authors liken to a Photoshop for robotics, still seems to have a significant learning curve, especially for casual consumers who just want to teach their robot to do something useful in the home.
 
@@
 
-As another example, robots that are capable of automated navigation may fail to observe human norms when navigating through human spaces.
+As another example, robots that are capable of automated navigation may fail to observe human norms when navigating through human spaces~\cite{lu-iros13}.
 The custom of staying to the right when passing in a hallway, for example, can cause confusion if a robot charges down the middle of the hall and does not yield to oncoming passerby.
 
-Endowing computers with common sense is still an active research area~\cite{mueller-cacm09}.
-Techniques for making use of common sense reasoning have not yet made it into the vocabulary of professional programmers.
+Endowing computers with common sense is still an active research area~\cite{mueller-cacm09}, as is making use of common sense knowledge in robotics~\cite{gupta-aaai04}.
 Yet these concepts will be taken for granted by end users of robots, who will expect their robots to obey common social norms as they perform their tasks.

### 2014-09-04 paper.tex
More refs and rewrote parts of intro section more formally

@@
 I never thought I would be a roboticist.
-In grad school, our robotics team spent months teaching little robot dogs to play soccer and solving other seemingly simple problems.
-Far better to work in the world of software, I figured, where the systems I built could make a real difference in people's lives.
-Software could automate repetitive tasks in computer use and free people from mundane work.
-So instead I developed expertise in human-computer interaction and intelligent user interfaces.
-My chosen field, end user programming, has the potential to enable millions of ordinary computer users to customize and adapt systems to their own needs.
+In grad school I focused on software agents -- {\em softbots}~\cite{etzioni-cacm94} -- that automated repetitive tasks and freed people from mundane work.
+I developed expertise in human-computer interaction and intelligent user interfaces.
+The field of end user programming, where I have focused most of my research, has the potential to enable millions of ordinary computer users to customize and adapt software systems to their own needs.
 
@@
 Even more, we are now seeing robots tackle tasks from our daily lives.
-Robotic vacuum cleaners are changing the way we clean our homes.
-Robotic lawnmowers keep the yard tidy without having to lift a finger.
-Nest's robotic thermostat learns your habits and keeps your home at the optimal temperature.
-Google's driverless car could change the way we commute.
+Robotic vacuum cleaners are changing the way we clean our homes\footnote{http://www.irobot.com/us/learn/home/roomba}.
+Robotic lawnmowers keep the yard tidy without having to lift a finger\footnote{http://www.robomow.com}.
+Nest's robotic thermostat learns your habits and keeps your home at the optimal temperature\footnote{http://www.nest.com}.
+Google's Self-Driving Car could change the way we commute.
 
@@
 
-Yet the holy grail of personal service robots remains elusive.
+One of the main applications for robotics is to create personal service robots which can assist humans in performing activities of daily living.
+This vision is particularly attractive as our population ages, and we wish to maintain or increase elders' quality of life despite the rapidly growing costs in the health-care sector~\cite{roy-wire00}.
+
+Yet the goal of creating personal service robots remains elusive.
 I believe there are three main reasons.
 First, there is a wide variety of tasks that people want service robots to perform, ranging from folding laundry, to finding lost car keys, to picking up the kids' toys, or doing the dishes and putting them away.
-Second, these tasks must be performed in a variety of environments, including cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
+Second, these tasks must be performed in a variety of environments, such as cluttered homes and kitchens with different affordances such as modern drawer pulls or faucet handles.
 Finally and most importantly, the robots themselves must be easy to operate by non-technical people.
@@
 On the other hand, Rethink Robotics' Baxter robot\footnote{http://www.rethinkrobotics.com/products/baxter/} can be easily programmed to perform a variety of tasks in light industrial settings, but its lack of mobility makes it less suitable for the home.
-As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists.
+As another example, Willow Garage's PR-2 robot has been programmed to perform a variety of household tasks, in a variety of environments, but can only be operated by highly skilled roboticists~\cite{ciocarlie-iser10}.
 
-Yet I believe the time is ripe for a breakthrough in personal service robotics, starting with robots that perform simpler tasks in relatively structured environments such as hospitals or hotels, and driven by better design and usable interfaces that enable them to be used by non-expert users.
+I believe the time is ripe for a breakthrough in personal service robotics, starting with robots that perform simpler tasks in relatively structured environments such as hospitals or hotels, and driven by better design and usable interfaces that enable them to be used by non-expert users.
 
@@
 Yet our experience with Robots for Humanity underscored how difficult it was for ordinary mortals to instruct robots what to do.
-Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door was very difficult for Henry to accomplish.
+Even using state-of-the-art robot visualization software developed at Willow Garage, simple teleoperation tasks such as opening a cabinet door were very difficult for Henry to accomplish.
 While Henry had only limited interaction capabilities, he was also strongly motivated because he has no alternatives.
@@
 
-Granted, designing teleoperation interfaces for motor-impaired users is deliberately trying to solve the hardest problem first.
-Yet while Henry's case highlights the challenges of creating usable assistive robotics interfaces, it also illustrates the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
-
+While Henry's case highlights the challenges of creating usable assistive robotics interfaces, it also illustrates the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
 With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
@@
 End users must select which behavior to call when, in order to make the robot do what they want.
-%Anyone who operates a robot, including consumers who purchase one for their home, will be programming the robot according to this definition.
+
+Even the seemingly-simple cleaning functionality of a Roomba vacuum is typically invoked deliberately as part of a larger sequence of housecleaning tasks.
+For example, users often reported first doing pre-cleaning tasks such as picking up clutter or setting down obstacles to constrain the robot's path before activating it~\cite{forlizzi-hri06}.
+Guided by knowledge of the robot's capabilities, these users are selecting when to invoke the robot's functionality in order to have the desired result.
+Robots with more functionality only increase the difficulty of this programming process.
 
@@
 \item {\bf Task/environment variance}: if the task to be performed is always the same, the programming task is easier than if the goal task changes significantly or if the environment in which the robot must operate is changing.
-The program must be generalized to handle this variance, which makes the programming process more difficult.
-% REF: something from the EUP literature? Andy Ko or Margaret Burnett?
+The program must be generalized to handle this variance, which makes the programming process more difficult~\cite{cypher-chi91,bogart-vlhcc08}.
 
@@
 These interfaces for robot programming will need to be simple enough for ordinary consumers to be able to pick up and use.
-Until users are able to instruct robots quickly and with low effort, personal robotics will be limited to niche markets.
+Until users are able to instruct robots quickly and with little effort, personal robotics will be limited to niche markets.
 Why is end user robot programming so challenging?
@@
 
-%Most robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
+Many robot programs can be simplified by only controlling one moving part at a time: first move the base, then raise the torso, then position the left arm, then position the right arm, finally open the cabinet.
 For many tasks this is a reasonable solution.
-However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands, or turning one's head to look where one is going.
+However there is a subset of human tasks for which we need multiple parts moving simultaneously, such as picking up a heavy or bulky object with two hands.
 
@@
 In order to deal with these various types of uncertainty, professional programmers learn to use probabilistic models~\cite{darwiche-cacm10} and craft fault-tolerant code~\cite{wright-cacm09} with frequent error checking.
-Systems such as Nguyen's ROS Commander~\cite{nguyen-icra13} enable non-roboticists to define and customize robot behaviors, handling error conditions.
+Systems such as Nguyen's ROS Commander~\cite{nguyen-icra13} enable non-roboticists to define and customize robot behaviors, including what to do under different error conditions.
 However, ROS Commander's programming interface, which the authors liken to a Photoshop for robotics, still seems to have a significant learning curve, especially for casual consumers who just want to teach their robot to do something useful in the home.
@@
 Reasoning about objects in the world requires some amount of common sense which has yet to be programmed consistently into robots.
-For example, a robot who has been asked to pick up and deliver a coffee cup may well select a grasp that involves dipping its fingers into the liquid and picking up the cup by its side wall.
-Although the grasp would be valid, this behavior would be judged somewhat undesireable by human observers.
+For example, a robot who has been asked to pick up and deliver a coffee cup will need to be taught to keep the cup upright during transit to avoid spilling its contents.
 

### 2014-09-04 paper.tex
Work work work

@@
 
+Work on robot programming by demonstration (e.g., Cakmak {\sl et al}~\cite{cakmak-rss10}) reduces the need to interact with 3D visualizations by having users directly manipulate the robot.
+Nevertheless, participants in their user study found the 3D environment difficult to navigate TBD!!!
+
+!!! Talk about Maya's work here on PBD with interactive Action Visualizations
+http://www.roboticsproceedings.org/rss10/p48.pdf
+
 %\section{7-DOF arms}

### 2014-09-10 paper.tex
Final revision submitted on Sept 4, 2014

--- 2014-09-04 paper.tex
@@
 \maketitle
-
-% Maya's Robot PBD
-% Strips style robot planning
 
@@
 While Henry's case highlights the challenges of creating usable assistive robotics interfaces, it also illustrates the enormous benefit to humanity of enabling disabled adults to regain some of their independence.
-With this goal in mind, personal service robots are more than a consumer toy, but a technology that could changes people's lives for the better.
+With this goal in mind, personal service robots are more than a consumer toy, but a technology that could change people's lives for the better.
 Why, then, are robots still so difficult to use?
@@
 
-\section{Robot programming}
+\section{Programming robots}
 
@@
 These are the primitive behaviors that have been programmed into the robot and can be activated by its user.
-End users must select which behavior to call when, in order to make the robot do what they want.
+End users must select which behavior to call, and when, in order to make the robot do what they want.
 
@@
 Many of the lessons learned from the fields of end user programming~\cite{dontcheva-nocode} and end user software engineering~\cite{ko-acm11} also apply to robot programming, though robots' physical embodiment leads to additional challenges.
-Based on my experience in both end user programming and robotics, I believe that the difficulty of the robot programming task boils down to several factors:
+Based on my experience in both end user programming and robotics, I would characterize robot programming tasks along a number of dimensions:
 
@@
 
-\subsection{A 3D world}
+\subsection{A 4D world}
 
@@
 
-Unlike traditional computer interfaces which are limited to two dimensions, robot control requires manipulating objects in a three-dimensional world.
+Unlike traditional computer interfaces which are limited to two dimensions, robot control requires manipulating objects in a three-dimensional world, with time providing a fourth dimension.
 Visualizing a 3D world on existing 2D displays requires the extra dimension to be projected down into two dimensions, giving rise to the popular ``first person shooter'' style of interface.
@@
 Additional camera controls are needed to pan/zoom around the space to overcome occlusion.
-These controls are often non-intuitive, even to seasoned computer users.
+These controls are often non-intuitive, even to seasoned computer users~\cite{ioannidou-jvlc09}.
+In addition, any 3D visualization shows only a single snapshot in time, requiring additional controls or visualizations to convey changes to the world over time.
 
-Occlusion presents a challenge because robots must perceive an object in order to manipulate it.
+%Occlusion presents a challenge because robots must perceive an object in order to manipulate it.
 Often times, the robot itself might occlude the object it is trying to manipulate!
@@
 While moving the arm into position to grasp the object, the arm will likely come between the head and the object, obscuring the robot's vision.
-
-One solution has been to mount additional cameras on different parts of the robot, such as on the robot's forearm or wrist, to get a better view on the object being picked up.
+Subsequently planning a path to move the object to a new position without colliding with anything in the environment is still an active research area~\cite{chitta-moveit}.
+%
+%One solution has been to mount additional cameras on different parts of the robot, such as on the robot's forearm or wrist, to get a better view on the object being picked up.
 However, making sense of multiple translated and rotated video streams can be confusing for those of us accustomed to perceiving the world only through head-mounted eye sensors!
-
-Typical input devices such as mice and touchscreens are also only good for interacting in two dimensions.
+%
+%Typical input devices such as mice and touchscreens are also only good for interacting in two dimensions.
 For example, tapping on a point in a camera image does not uniquely identify a point in 3D space, but a line of possible points.
@@
 Current interfaces, such as the RViz display shown in Figure~\ref{rviz}, require complex camera manipulation and precise 3D positioning in order to uniquely specify the desired location.
+
+Work on robot programming by demonstration (e.g., Alexandrova {\sl et al}~\cite{cakmak-rss14}) reduces the need to interact with 3D visualizations by having users directly manipulate the robot.
+In their system, users demonstrate a desired robot behavior once by directly manipulating the robot's arms, and then uses a graphical 3D tool to generalize and modify the learned behavior.
+Yet participants in their user study found it difficult to interact with the 3D environment.
 
@@
 
-The field of end user programming~\cite{lieberman-yourwish,dontcheva-nocode} has developed a variety of solutions to enable users of software systems to create or modify software artifacts without having to learn the intricacies of programming.
+The field of end user programming (EUP)~\cite{lieberman-yourwish,dontcheva-nocode} has developed a variety of solutions to enable users of software systems to create or modify software artifacts without having to learn the intricacies of programming.
 
-The key insight for these EUP systems is to reduce the complexity of a programming task through simplification.
-The three primary methods include (a) visual programming: simplifying or eliminating tricky syntax through graphical representations; (b) domain-specific languages: limiting concepts to only those necessary in a particular domain of use, such as spreadsheet formulas; and (c) programming by demonstration or example: letting users focus on concrete examples rather than difficult abstractions.
+The key insight for these EUP systems is to reduce the complexity of a programming task through simplification, trading off expressiveness for ease of use.
+Three primary methods include (a) visual programming: simplifying or eliminating tricky syntax through graphical representations; (b) domain-specific languages: limiting concepts to only those necessary in a particular domain of use, such as spreadsheet formulas; and (c) programming by demonstration or example: letting users focus on concrete examples rather than difficult abstractions.
 
@@
 This abstraction can be difficult for roboticists to accept, who have become accustomed to having fine-grained control over robot behavior and error handling.
-Yet abstracting away these details is going to be critical for end users to be able to develop a working understanding of a robot's functionality.
+Yet abstracting away these details is going to be critical for end users to be able to define useful robot behaviors.
 
@@
 We made use of these autonomous functions to develop a system at Willow Garage called Continuous Ops (short for ``Continuous Operations'') that enabled casual users to control PR-2 robots through a web interface (Figure~\ref{contops}).
-The interface was inspired by the head of Willow Garage, who noted that even though he had a lab full of robots, he himself wasn't able to show off any of the functionality of his robots to visitors without a roboticist present to operate them.
+The interface was inspired by the head of Willow Garage, who noted that even though he had a lab full of robots, he himself was not able to show off any of the functionality of his robots to visitors without a roboticist present to operate them.
 
-In response, we developed a web interface, accessible from standard laptops or tablets, that provided the ability to tell a PR-2 robot to perform a variety of tasks.
+In response, we developed a web interface, accessible from standard laptops or tablets, that provided the ability to command a PR-2 robot to perform a variety of tasks.
 It could unplug itself from the power supply, navigate autonomously to anywhere in the buliding, position itself in front of a table, automatically identify graspable objects on that table, and pick up an object.
@@
 
-Another way to simplify the programming problem for end users is to take advantage of {\em shared autonomy}.
+Another way to simplify the programming problem for end users is to take advantage of {\em shared autonomy}~\cite{leeper-hri12,pitzer-icra11}.
 The idea behind shared autonomy is to fall back on teleoperator assistance when autonomy fails, as it will invariably do for the near future.
@@
 With a human expert on call to get a robot out of situations it was not programmed to handle, we can simplify the programming problem for end users so that they do not have to specify error recovery behaviors for every single corner case.
-
 As a stepping-stone towards full autonomy, this approach also enables the testing of end user programmable systems without needing a fully reliable set of autonomous primitives in place to support them.
@@
 This approach worked well during a multi-day deployment of a delivery system at Willow Garage.
-Users could request delivery of chocolates or snacks to their office, and the system would autonomously retrieve those items and deliver them to their office door.
+Users could request delivery of chocolates or snacks to their office, and a robot would autonomously retrieve those items and deliver them to their office door.
 In the rare cases that the autonomous behaviors failed, several roboticists were notified by email and played the role of teleoperators.
@@
 
-Finally, video gaming technology is driving a new crop of input and output devices that make it easier to interact with and perceive the 3D world around us.
+Finally, video gaming technology is driving a new crop of input and output devices that make it easier to interact with and perceive the 3D world around us~\cite{goth-cacm11}.
 These devices will make it easier for robots to perceive the world in three dimensions, and for people to instruct those robots using more natural gestures.
@@
 These new devices could enable a new generation of robot control interfaces that make interacting with robots more natural and intuitive.
-For example, David Gossow at Willow Garage developed a research prototype called the PR-2 Surrogate which provided a fully-immersive teleoperation interface for the PR-2 robot.
+For example, David Gossow at Willow Garage developed a research prototype called the PR-2 Surrogate which provided a fully-immersive teleoperation interface for the PR-2 robot\footnote{https://www.youtube.com/watch?v=H0EoEyvTmiY}.
 It used a Razer Hydra control to track your hand positions, which were then mapped directly to the robot's hand positions.
@@
 Personal service robots have an enormous potential to do good for humanity by enabling people to live more independent lives and free them from mundane housekeeping chores.
-However, controlling these robots boils down to a small matter of programming.
+However, controlling these robots boils down to a small matter of programming~\cite{nardi-smop}.
 In this article I have explained why programming robots is inherently difficult, even for skilled programmers.
